---
title: "Muti epitope maps -- ferrets"
output: html_notebook
---

** This notebook modifies v0.0 so that we can use a single function to test differnt antigen maps and immunodominance levels**

Question --

When we use titers to map a multi-epitope pathogen, how does the inferred "strain" location relate to the locations/distances of individual epitopes?

**Workflow:**

1. Simulate a three-epitope pathogen with 5 antigens of each epitope represented on a 2D map.
2. Simulate ferret antibody repertoires to the epitope.
3. Calculate each ferret's titer to each strain to simulate a sero panel.
4. From the sero panel, infer the "strain map".
5. Repeat under conditions of even, skewed, and variable immunodominance

Qs:

* How do the inferred strain positions align with individual epitope positions?
* How does the relationship between strain and epitope position change as immunodominance becomes skewed?
* Can we infer the map consistently if ID varies between individuals?


```{r}
source('multi_epitope_code.R')
source('../code.R')
library('foreach')
reset() ## clear workspace and load packages


n_dim = 2
n_antigens = 5
n_ab = 0
n_chains = 4
n_iter = 5000 # half will be burn in
n_epitopes = 3
```

## Generate antigen coordinates. Each antigen consists of 3 epitopes.
```{r}
set.seed(13)
antigen_coords <- tibble(epitope = rep(c(1:n_epitopes), each = n_antigens),
                         antigen = rep(1:n_antigens, n_epitopes),
                         kind = 'antigen',
                         c1 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c2 = generate_ag_coords_random(n_antigens, n_epitopes)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))


## Standardize so that Ag1 epitope 1 is at the origin and Ag2 epitope 1 is on the x-axis
shifted_coords <-  standardize_coordinates(coord_mat = antigen_coords %>% select(starts_with('c')) %>% as.matrix(), ag1_row = 1, ag2_row = 2) %>%
  as_tibble() %>%
  set_names(c('c1', 'c2'))

antigen_coords$c1 = shifted_coords$c1
antigen_coords$c2 = shifted_coords$c2
```

## Plot the coordinates
```{r}
antigen_coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), alpha = .2)
ggsave('sample_ag_plot.png', )
```


```{r}
even_immunodominance <- infer_ferret_map(antigen_coords, 
                relative_concentrations = c(1, 1, 1), 
                n_epitopes = n_epitopes, 
                n_antigens = n_antigens, 
                n_abs_per_serum = 500, 
                plotdir = 'even_immunodominance', 
                sigma = .1)
```


```{r}
skewed_immunodominance <- infer_ferret_map(antigen_coords, 
                relative_concentrations = c(1+2/3, 1, 1/3), 
                n_epitopes = n_epitopes, 
                n_antigens = n_antigens, 
                n_abs_per_serum = 500, 
                plotdir = 'skewed_immunodominance', 
                sigma = .1)
```


```{r}
E1_complete_immunodominance <- infer_ferret_map(antigen_coords, 
                relative_concentrations = c(3, 0, 0), 
                n_epitopes = n_epitopes, 
                n_antigens = n_antigens, 
                n_abs_per_serum = 500, 
                plotdir = 'all_or_nothing_immunodominance', 
                sigma = .1)
```



## When immunodominance is even, differences in titer are smaller than differences between epitopes

* The inferred map is compressed (usually w.r.t. axis 2), relative to the centroid of the polygon defined by the corresponding epitopes.

```{r}
inferred_map <- extract_summary_coords(even_immunodominance$stan_fits) %>%
  mutate(id = as.factor(id)) %>%
  mutate(c2 = ifelse(chain %in% c(1,3,4), -c2, c2))


  ag1_ag2_centroids = antigen_coords %>%
    group_by(antigen) %>%
    summarise(epitope = NA,
              antigen = unique(antigen), 
              kind = 'antigen',
              c1 = mean(c1),
              c2 = mean(c2))
  
  true_coords = standardize_coordinate_df(coord_df = bind_rows(ag1_ag2_centroids, antigen_coords), # Standardize to Ag1, Ag2 centroids
                                          ag1_row = 1, ag2_row = 2) %>%
    filter(!is.na(epitope)) %>% # Remove ag1, ag2 centroids
    group_by(antigen) %>%
    mutate(centroid1 = mean(c1),
           centroid2 = mean(c2))

ggplot() +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), data = true_coords, alpha = .05) +
  geom_point(aes(x = centroid1, y = centroid2, color = antigen), data = true_coords, pch = 1) +
  geom_point(aes(x = c1, y = c2, color = id, shape = kind), data = inferred_map)
```




```{r}
plot_inferred_original_map(antigen_coords,
                           even_immunodominance$stan_fits,
                           flip_these_chains_over_x = c(1,3,4)) + 
  ggtitle('even') + 
  theme(legend.position = 'none') + 
  facet_grid(allele~chain, labeller = label_both) +
  geom_hline(aes(yintercept = 0), lwd = .5, alpha = .5)+
  geom_vline(aes(xintercept = 0), lwd = .5, alpha = .5)
```

```{r}
plot_inferred_original_map(antigen_coords,
                           even_immunodominance$stan_fits,
                           flip_these_chains_over_x = c(1,3,4)) + 
  ggtitle('even') + 
  theme(legend.position = 'none') +
  facet_wrap(.~chain, labeller = label_both)
```

## This compression persists, but becomes less severe when immunodominance is skewed.


```{r}
plot_inferred_original_map(antigen_coords,
                           skewed_immunodominance$stan_fits,
                           flip_these_chains_over_x = c(1,3)) +
  ggtitle('skewed') + theme(legend.position = 'none')  + 
  facet_grid(allele~chain, labeller = label_both)
```

```{r}
plot_inferred_original_map(antigen_coords,
                           skewed_immunodominance$stan_fits,
                           flip_these_chains_over_x = c(1,3)) +
  ggtitle('skewed') + theme(legend.position = 'none')  + 
  facet_wrap(.~chain, labeller = label_both)
```




## When a single epitope is immunodominant, we infer its coordinates exactly.

```{r}
plot_inferred_original_map(antigen_coords,
                           E1_complete_immunodominance$stan_fits,
                           flip_these_chains_over_x = 1:3) + 
  ggtitle('all or nothing') + theme(legend.position = 'none')  + 
  facet_grid(allele~chain, labeller = label_both) 
```

```{r}
plot_inferred_original_map(antigen_coords,
                           E1_complete_immunodominance$stan_fits,
                           flip_these_chains_over_x = 1:3) + 
  ggtitle('all or nothing') + theme(legend.position = 'none')  + 
  facet_wrap(.~chain, labeller = label_both) 
```


```{r}
cowplot::plot_grid(
cowplot::get_legend(even_immunodominance$inferred_map_plot)
)
```



```{r}
E1_complete_immunodominance$ag_ab_coords %>%
  ggplot() +
  geom_point(aes(x = c1_Ab, y = c2_Ab, color = epitope), pch = 3) +
  geom_point(aes(x = c1_Ag, y = c2_Ag, fill = epitope), pch = 21) +
  geom_point(aes(x = c1, y = c2, fill = epitope), data = antigen_coords, pch = 22) +
  guides(col = guide_legend(title = 'antibodies'))
```

