---
title: "Hosts differ in immunodominance"
output: html_notebook
---

Test map inference when hosts differ in immunodominance.


```{r}
source('multi_epitope_code.R')
source('../code.R')
library('foreach')
reset() ## clear workspace and load packages


n_dim = 2
n_ag = 5
n_ab = 0
n_chains = 4
n_iter = 5000 # half will be burn in
n_epitopes = 3
```

## Generate antigen coordinates. Each antigen consists of 3 epitopes.
```{r}
set.seed(13)
antigen_coords <- tibble(epitope = rep(c(1:n_epitopes), each = n_ag),
                         antigen = rep(1:n_ag, n_epitopes),
                         kind = 'antigen',
                         c1 = generate_ag_coords_random(n_ag, n_epitopes),
                         c2 = generate_ag_coords_random(n_ag, n_epitopes)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))


## Standardize so that Ag1 epitope 1 is at the origin and Ag2 epitope 1 is on the x-axis
shifted_coords <-  standardize_coordinates(coord_mat = antigen_coords %>% select(starts_with('c')) %>% as.matrix(), ag1_row = 1, ag2_row = 2) %>%
  as_tibble() %>%
  set_names(c('c1', 'c2'))

antigen_coords$c1 = shifted_coords$c1
antigen_coords$c2 = shifted_coords$c2
```

## Plot the coordinates
```{r}
antigen_coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), alpha = .2)
```

## Generate a set of hypothetical human sera. Initially:

* Assume each person is infected with just one strain
* Assume complete immunodominance of one epitope in each individual
* Generate at least 3 "sera" containing 1000 Abs each, for each individual.


To generate a serum:

* Assume antibody coordinates are normally distributed about the coordinates of the infecting strain's epitopes, with sigma = 0.1
* Assign antibodies to epitopes in proporiton to the epitope's host-specific imunodominance. Here, all Abs will be assigned to a single epitope.

```{r}
## Generate a data frame with serum metadata - which strain are Abs specific to and which epitope is immunodominant?
serum_catalog <- expand_grid(infecting_strain = 1:n_antigens,
                        dominant_epitope = 1:n_epitopes,
                        replicate = 1:10 # Do 10 replicates for each possible comination
                        ) %>%
  mutate(serum_id = 1:nrow(.))
```


```{r include=FALSE}
## Generate Ab populaitons for each serum.

serum_list <- foreach(this.infecting.strain = serum_catalog$infecting_strain,
        this.dominant.epitope = serum_catalog$dominant_epitope,
        serum.id = serum_catalog$serum_id) %do% {
          
          n_epitopes = nrow(unique(antigen_coords[,'epitope']))
          immunodominance_vec = vector(mode = 'numeric', length = n_epitopes)
          immunodominance_vec[this.dominant.epitope] = n_epitopes
          this_epitope_coords <- antigen_coords %>% 
            filter(antigen==this.infecting.strain & epitope == this.dominant.epitope)
          stopifnot(nrow(this_epitope_coords) == 1)
          generate_one_human_serum(native_epitope_coords = this_epitope_coords,
                                   n_epitopes = n_epitopes,
                                   total_ab = 1000,
                                   epitope_immunodominance = 1,
                                   strain_dominance = c(1),
                                   sigma = .1) %>%
            mutate(serum_id = serum.id)
        } 
```

## Generate a titer map 

* Calculate the titer of each serum to each antigen.

```{r}
titer_map <- matrix(NA, nrow = length(serum_list), ncol = n_antigens, dimnames = list(paste0('serum', serum_catalog$serum_id), paste0('antigen', 1:n_antigens)))
this.row = 0

foreach(serum = serum_list) %do%{
  this.row = this.row+1
  # Extract serum Ab coords as a list of matrices, one for each eptiope
  this_ab_position_list = lapply(1:n_epitopes, function(this.epitope){
    serum %>% 
      filter(epitope == this.epitope) %>%
      select(matches('c\\dAb')) %>% 
      as.matrix()
  })
for(this.antigen in 1:n_antigens){
    # Extract antigen coords as a list of matrices, one for each epitope
    this_antigen_list <- lapply(1:n_epitopes, function(this.epitope){
      antigen_coords %>% 
        filter(epitope == this.epitope & antigen == this.antigen) %>%
        select(starts_with('c')) %>%
        as.matrix()
    })
    # solve for titer  
    #cat(sprintf('This serum is %s, this antigen is %s', this.row, this.antigen))
    titer_map[this.row, this.antigen] = solve_for_titer_multi_epitope(ab_position_list = this_ab_position_list, 
                                                                       relative_concentrations = c(1,1,1), 
                                                                       ag_list = this_antigen_list, 
                                                                       alpha = .25, 
                                                                       r = 7)
  }
}

titer_map
```

```{r}
titer_df <- titer_map %>% as_tibble() %>%
  bind_cols(serum_catalog) %>%
  pivot_longer(starts_with('antigen'), names_to = c('test_antigen'), values_to = 'titer', names_prefix = 'antigen') %>%
  group_by(test_antigen) %>%
  mutate(titer_distance = max(log2(titer/10))-log2(titer/10))
```


```{r}
titer_df %>%
  ggplot() +
  geom_tile(aes(x = infecting_strain, y = test_antigen, fill = titer, group = serum_id))+
  facet_wrap(.~dominant_epitope, labeller = label_both) +
  scale_fill_viridis_c(direction = -1, option = 'B')
```

```{r}
titer_df %>%
  ggplot() +
  geom_tile(aes(x = infecting_strain, y = test_antigen, fill = titer_distance, group = serum_id))+
  facet_wrap(.~dominant_epitope, labeller = label_both) +
  scale_fill_viridis_c(direction = -1, option = 'B')
```
```{r}
## Compare the actual distances between epitopes to the calculated titer distances
get_one_distance <- function(this.epitope, antigen1, antigen2, antigen_coords){
  vec1 <- antigen_coords %>%
    filter(epitope == this.epitope) %>%
    filter(antigen == antigen1) %>%
    select(matches('c\\d')) %>%
    as.vector()
  
  vec2 <- antigen_coords %>% 
    filter(epitope == this.epitope) %>%
    filter(antigen == antigen2) %>%
    select(matches('c\\d')) %>%
    as.vector()
  
  stopifnot(length(vec1) == length(vec2))
  stopifnot(length(vec1) == ncol(select(antigen_coords, matches('c\\d'))))
  
  sqrt(sum( (vec1 - vec2)^2 ))
}


compare_distances <- expand_grid(dominant_epitope = 1:3,
            infecting_strain = 1:5,
            test_antigen = 1:5) %>%
  rowwise() %>%
  mutate(coord_distance = get_one_distance(this.epitope = dominant_epitope, antigen1 = infecting_strain, antigen2 = test_antigen, antigen_coords)) %>%
  merge(titer_df, all.x = TRUE)
```

```{r}
## Check that there is a clear relationship between titer and coordinate distances
compare_distances %>%
  ggplot() +
  geom_point(aes(x = (titer_distance), y = (coord_distance), color = infecting_strain, fill = test_antigen), pch = 23) +
  facet_wrap(.~dominant_epitope, labeller = label_both) +
  scale_color_viridis_c(aesthetics = c('color', 'fill'))
```

# Now, for various combinations of immunodominance, infer the maps:
```{r}
## Wrapper function to extract distance matrix for input into stan
get_titer_distmat <- function(valid_serum_ids,
                              titer_df){
  these.titers = titer_df %>%
    filter(serum_id %in% unique(valid_serum_ids)) 
  
  antigens = as.numeric(unique(these.titers$test_antigen))
  sera = as.numeric(unique(these.titers$serum_id))
  #cat(print(antigens))
  #cat(print(sera))
  
  distmat = matrix(NA, length(antigens), length(sera), dimnames = list(paste0('ag_', antigens), paste0('serum_', sera)))
  #cat(print(distmat))
  
  for(aa in 1:length(antigens)){
    for(ss in 1:length(sera)){
      this.dist = filter(these.titers, test_antigen == antigens[aa] & serum_id == sera[ss]) %>%
        pull(titer_distance)
      #print(this.dist)
      distmat[aa,ss] = this.dist
    }
  }
  distmat
}
```


```{r}
plot_one_epitope_fits <- function(include.epitopes, 
                                  these.fits,
                                  flip_these_chains # Which chains to reflect over x?
                                  ){
  true_coords = standardize_coordinate_df(coord_df = antigen_coords %>% filter(epitope %in% include.epitopes), ag1_row = 1, ag2_row = 2)
ggplot() +
  geom_point(data = true_coords, 
             aes(x = c1, y = c2, fill = antigen), pch = 23, size = 3, color = 'black', alpha = .3) +
  geom_point(data = extract_summary_coords(these.fits) %>%
               mutate(c2 = ifelse(chain %in% flip_these_chains, -c2, c2)),
             aes(x = c1, y = c2, color = id, shape = kind)) +
  facet_wrap(.~chain, labeller = label_both) +
  scale_shape_manual(values = c(3,1, 4))
}
```


```{r}
plot_log_posteriors <- function(stanfit){
  this.fit = rstan::extract(stanfit,permuted=F)
print(dimnames(this.fit))
this.fit[,,'lp__'] %>%
  as_tibble() %>%
  mutate(iter = 1:nrow(.)) %>%
  pivot_longer(-iter, names_to = 'chain', values_to = 'log_posterior') %>%
  ggplot() +
  geom_point(aes(x = iter, y = log_posterior, color = chain))
}
```

```{r}
downsample <- function(x, n_out){
  keep = sample(x, size=n_out, replace = FALSE)
  x %in% keep
}
```











## 1.1 Infer map for E1 sera only. We should recover the E1 coords exactly

```{r}
valid_serum_ids = serum_catalog %>% filter(dominant_epitope == 1) %>% pull(serum_id)
E1_only_fits <- fit_stan_MDS(observed_distances = get_titer_distmat(valid_serum_ids, titer_df = titer_df), 
                       n_antigens = n_antigens, 
                       n_sera = length(valid_serum_ids), 
                       n_dim = n_dim, 
                       chains = n_chains,
                       niter = n_iter)
```

```{r}
plot_epitope_fits(this.epitope = 1, 
                      E1_only_fits,
                      flip_these_chains = 2:4)
```




## 1.2 Infer map for E2 sera only. We should recover the E1 coords exactly

```{r}
valid_serum_ids = serum_catalog %>% filter(dominant_epitope == 2) %>% pull(serum_id)
E2_only_fits <- fit_stan_MDS(observed_distances = get_titer_distmat(valid_serum_ids, titer_df = titer_df), 
                       n_antigens = n_antigens, 
                       n_sera = length(valid_serum_ids), 
                       n_dim = n_dim, 
                       chains = n_chains,
                       niter = n_iter)
```

```{r}
plot_epitope_fits(this.epitope = 2, 
                      E2_only_fits,
                      flip_these_chains = c(1))
```
This is strange... Why do chains 3 and 4 converge to an estimate for antigen 5 that's reflected in x?
If they're getting stuck at a local optimum, then we should make sure that the log posterior density of those chains is lower than the log posterior of the better chains.
The plot below verifies that this is true.

```{r}
plot_log_posteriors(stanfit = E2_only_fits)
```






## 1.3 Infer map for E3 sera only. We should recover the E1 coords exactly

```{r}
valid_serum_ids = serum_catalog %>% filter(dominant_epitope == 3) %>% pull(serum_id)
E3_only_fits <- fit_stan_MDS(observed_distances = get_titer_distmat(valid_serum_ids, titer_df = titer_df), 
                       n_antigens = n_antigens, 
                       n_sera = length(valid_serum_ids), 
                       n_dim = n_dim, 
                       chains = n_chains,
                       niter = n_iter)
```

```{r}
plot_epitope_fits(this.epitope = 3, 
                      E3_only_fits,
                      flip_these_chains = 3)
```

```{r}
plot_log_posteriors(E3_only_fits)
```







# Ok, now see what happens when we try to infer titers using sera that are specific to different epitopes.


## Let's start by doing an even mix of sera that are specific to E1 and E2
```{r}
valid_serum_ids = serum_catalog %>% 
  filter(dominant_epitope != 3) %>% # Exclude sera specific to E3
  group_by(dominant_epitope) %>%
  filter(downsample(1:n(), n_out = 25)) %>% ## Keep 25 sera specific to each epitope
  pull(serum_id)

E1_E2_blanced <- fit_stan_MDS(observed_distances = get_titer_distmat(valid_serum_ids, titer_df = titer_df), 
                       n_antigens = n_antigens, 
                       n_sera = length(valid_serum_ids), 
                       n_dim = n_dim, 
                       chains = n_chains,
                       niter = n_iter)
```

```{r}
plot_two_epitope_fits <- function(include.epitopes, 
                                  these.fits,
                                  flip_these_chains, # Which chains to reflect over x?
                                  valid_serum_ids,
                                  serum_catalog
                                  ){
  true_coords = standardize_coordinate_df(coord_df = antigen_coords %>% filter(epitope %in% include.epitopes), ag1_row = 1, ag2_row = 2)%>%
    mutate(kind=paste0('epitope', epitope))%>%
    mutate(size =2,
           alpha =.7,
           antigen = as.numeric(antigen)) %>%
    expand_grid(chain = 1:4)
  
  inferred_coords=extract_summary_coords(these.fits) %>%
    mutate(c2 = ifelse(chain %in% flip_these_chains, -c2, c2),
           kind = paste0('inferred_', kind)) %>%
    arrange(kind, id) %>%
    mutate(serum_id = NA)
  inferred_coords$serum_id[inferred_coords$kind =='inferred_serum'] = valid_serum_ids
  inferred_coords =merge(inferred_coords,
                         serum_catalog,by ='serum_id',
                         all.x = T,all.y = F)%>%
    arrange(kind, id) %>%
             mutate(size = ifelse(kind != 'inferred_serum', 1.2, 1), 
                    alpha = 1) %>%
             mutate(antigen = ifelse(grepl('serum', kind), infecting_strain, id)) %>%
    filter(kind != 'inferred_serum')
  
  bind_rows(true_coords%>% select(c1, c2, kind, antigen, size, alpha, chain),
            inferred_coords %>% select(c1, c2, kind, antigen, size, alpha, chain)) %>%
    mutate(antigen =as.factor(antigen)) %>%
    ggplot()+
    geom_point(aes(x = c1, y = c2, color = antigen, shape = kind, size =size, alpha =alpha))+
  scale_shape_manual(values =c(16,17, 2, 5, 3))+
    scale_size(range =c(2,3), guide = FALSE)+
    scale_alpha(range =c(.5, 1), guide =FALSE)+
    facet_wrap(.~chain, labeller = label_both)
}
```


```{r}
plot_two_epitope_fits(include.epitopes = c(1,2), these.fits = E1_E2_blanced, flip_these_chains = NULL, valid_serum_ids = valid_serum_ids, serum_catalog)
```




