---
title: "v0.2_LS"
author: "Katie Gostic"
date: "1/24/2022"
output: html_document
---

## This is a version of the script `` that runs in 3D

* Visulation is not really attempted, this is just to demonstrate that the code works
## Test in 3D

```{r setup, include=FALSE}
rm(list = ls())
library(tidyr)
filter <- dplyr::filter
extract <- tidyr::extract
select <- dplyr::select
source('../code.R')
```

```{r cars}
coords_3D <- generate_map_coordinates(n_dim = 3, n_antigens = 5, n_sera = 6, map_range = c(-10, 10))
coords_3D
```

```{r}
# Like most of the functions in thei script, `get_ab_ag_distances` is a function defined in `../code.R`
distmat3D <- get_ab_ag_distances(ab_coords = filter(coords_3D, kind == 'serum') %>% coord_tibble_to_matrix(),
                               ag_coords = filter(coords_3D, kind == 'antigen') %>% coord_tibble_to_matrix())
```


```{r}
# Run the optimization algothirm
fit_list_3D <- fit_MDS_least_squares(observed_distances = distmat3D, # Matrix of observed distances (n_antigens x n_sera)
                                  n_antigens = nrow(distmat3D), # Scalar, n antigens. Must match nrow(observed_distances).
                                  n_sera = ncol(distmat3D),  # Scalar, n sera. Must match ncol(observed_distances)
                                  n_dim = 3 # Dimensions in the map
                                  )
```



## Extract the best fit

```{r}
best_fit <- get_fits_df(fit_list_3D$best_fit)%>%
  mutate(kind = ifelse(kind == 'ag', 'antigen', 'serum'))
```


## Verify that inferred distances match the original
```{r}
cat(('Inferred distance matrix:\n'))
best_fit_distmat <- get_ab_ag_distances(ab_coords = best_fit %>% filter(kind == 'serum') %>% select(starts_with('c')) %>% as.matrix(),
                                        ag_coords = best_fit %>% filter(kind == 'antigen') %>% select(starts_with('c')) %>% as.matrix())
best_fit_distmat
```

```{r}
cat(('Actual distance matrix:\n'))
distmat3D
```
## Tranform to match the original basis

--> The isometric tranformation function I wrote only works in 2D.

I may revist this at a later date to try to generalize to higher dimensions.


