---
title: "Bayesian MDS in Stan"
output: html_notebook
---

Write and test a Stan model

```{r setup, include=FALSE}
rm(list = ls())
library(tidyr)
filter <- dplyr::filter
extract <- tidyr::extract
select <- dplyr::select
source('../code.R')
```


```{r cars}
n_antigens = 5
n_sera = 6
n_dim = 2
coords <- generate_map_coordinates(n_dim = n_dim, 
                                   n_antigens = n_antigens, 
                                   n_sera = n_sera, 
                                   map_range = c(-10, 10))
coords
```

## Plot the synthetic coordinates
```{r}
original_coord_plot <- coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = kind))
```

## Get the distances between coordinates

* We only observe the distances between antigen (ag) and antibody/serum (ab), not between ag-ag or ab-ab.
* Assume all ag-ab distances are observed without error

```{r}
# Like most of the functions in thei script, `get_ab_ag_distances` is a function defined in `../code.R`
distmat <- get_ab_ag_distances(ab_coords = filter(coords, kind == 'serum') %>% coord_tibble_to_matrix(),
                               ag_coords = filter(coords, kind == 'antigen') %>% coord_tibble_to_matrix())
```

## Infer map using basic stan model


To get this working, let's initalize using the exact solution 
```{r}
  initfun <- function(){
    list(sigma = .05,
         strain_coords = (coords %>% filter(kind == 'antigen') %>% select(starts_with('c')) %>% as.matrix),
         serum_coords = (coords %>% filter(kind == 'serum') %>% select(starts_with('c')) %>% as.matrix)
         )
    }
```


```{r}
fit_stan_MDS <- function(
  mod = 'MDS.stan',
  observed_distances, # n_antigen x n_antibody matrix of distances
  n_antigens, # integer
  n_sera, # integer
  n_dim, # integer
  chains = 3, # Number of MCMC chains to run
  cores = parallel::detectCores(logical = F), # For the cluster
  ...
) {
  library(rstan)
  
  stopifnot(nrow(observed_distances) == n_antigens)
  stopifnot(ncol(observed_distances) == n_sera)
  
  model <- stan_model('MDS.stan')
  
  model_input_data <- list(
    n_strains = n_antigens,
    n_sera = n_sera,
    n_dim = n_dim,
    observed_distances = observed_distances
  )
  
  fit <- sampling(
    model, model_input_data, chains = chains, cores = cores, 
    init = initfun,
    iter = 5000,
  #  control = list(adapt_delta = 0.8,
  #                  max_treedepth = 13),
    ...
  )
}

```

```{r}
fits <- fit_stan_MDS(observed_distances = distmat, 
                     n_antigens = n_antigens, 
                     n_sera = n_sera, 
                     n_dim = n_dim, 
                     chains = 1)
```

## Visualize trace plots
```{r}
rstan::traceplot(fits, pars = names(fits))
```

## Visualize the pairs plots
```{r}
pairs(fits, pars = names(fits)[1:5])
```

## Check the distance matrix
```{r}
raw_fits <- rstan::extract(fits)

long_antigen_coords <- lapply(1:n_antigens, function(ll) raw_fits$antigen_coords[,ll,] %>%
                                                      magrittr::set_colnames(c('c1', 'c2')) %>%
                           as.tibble() %>%
                           mutate(id = ll,
                                  iter = 1:nrow(.),
                                  kind = 'antigen')) %>%
                    bind_rows()

long_serum_coords <- lapply(1:n_sera, function(ll) raw_fits$serum_coords[,ll,] %>%
                                                      magrittr::set_colnames(c('c1', 'c2')) %>%
                           as.tibble() %>%
                           mutate(id = ll,
                                  iter = 1:nrow(.),
                                  kind = 'serum')) %>%
                    bind_rows()

summary_coords <- bind_rows(long_antigen_coords,
          long_serum_coords) %>%
  group_by(id, kind) %>%
  summarise(c1.10 = quantile(c1, .1),
            c1 = median(c1),
            c1.90 = quantile(c1, .9),
            c2.10 = quantile(c2, .1),
            c2 = median(c2),
            c2.90 = quantile(c2, .9)) %>%
  ungroup()
```

## Plot
```{r}
cowplot::plot_grid(
summary_coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = kind)) +
  ggtitle('stan') +
  theme(legend.position = 'none'),
original_coord_plot +
  ggtitle('original') +
  theme(legend.position = 'none')
)
```
 
 Ok, this looks reasonable, but the MDS coords are shifted like crazy.

## Get the distance matrix and compare to the original.
## If the distances are reasonable, then the inference could be working.
```{r}
cat('Inferred distances (median):\n')
get_ab_ag_distances(ab_coords = summary_coords %>% filter(kind == 'serum') %>% select(matches('c\\d$')) %>% as.matrix(), 
                    ag_coords = summary_coords %>% filter(kind == 'antigen') %>% select(matches('c\\d$')) %>% as.matrix()
)

cat('\nActual distances:\n')
distmat
```


## Ok, this is working, but I think it makes sense to fix the first ag coordinate, and the direction of vec(ag_1, ag_2) so that we don't have to wory about the isometry issue so much.