---
title: "Dimension reduction sandbox"
output: html_notebook
---

Before running PCR on simulated titers, try running it on an idealized distnace matrix that we expect to reduce to 1D. 

## Write out the 1D model

* $z_{ik}$ titer between strain i and serum k
* $x_i$ is the D1 coordinate of strain i
* c is a constant scaling the decrease in log2 titer with distance


$z_{ik} = z^{max}_k - c(\sqrt(x_i-x_k)^2)$

## Simulate 1d titers
```{r}
z_max = 10
get_one_titer_1D <- function(z_max, xi, xk, c = 1){
  z_max - sqrt((xi-xk)^2)
}

titer_df_1D <- expand_grid(x_strain = 1:10,
            x_serum = 1:10) %>%
  rowwise() %>%
  mutate(titer = get_one_titer_1D(10, x_strain, x_serum))

titer_mat_1D = matrix(titer_df$titer, nrow = 10, ncol = 10, byrow = F, dimnames = list(paste0('strain', 1:10), paste0('serum', 1:10)))
```

## Perform PCA on 1D titers
```{r}
PCA_1D <- prcomp(x = titer_mat_1D, scale = TRUE, center = TRUE)
summary(PCA_1D)
```

Even in the 1D model, we need 2d to represent the titer map.
Presumably, d1 represents the linear decrease in titer with distance, and d2 represents the peak titer for each dimension?

## Visualize
```{r}
ggbiplot::ggbiplot(PCA_1D, labels = rownames(titer_mat_1D)) +
  ylim(c(-2, 2)) + xlim(c(-2,2))
```

# Extend to 2D

If the above interpretation is true, then if we define the titer model in 2D, we should need 3 dimensions to represent the titer map.


## 2D model:

$z_{ik} = z^{max}_k - c(\sqrt(x_{1i}-x_{1k})^2+(x_{2i}-x_{2k})^2)$


## Simulate 2D titers
```{r}
get_one_titer <- function(z_max, xi, xk, ndim = 1, c = 1){
  stopifnot(length(xi) == length(xk))
 # cat(sprintf('xi is %s\n', xi))
  stopifnot(length(xi) == ndim)
  this.dist = get_euclidean_distance(xi, xk)
  #cat(sprintf('this distnace is %s\n', this.dist))
  z_max - this.dist
}

titer_df_2D <- expand_grid(x1_strain = 1:10,
                           x2_strain = 1:10,
            x1_serum = 1:10,
            x2_serum = 1:10) %>%
  group_by(x1_strain, x2_strain) %>%
  mutate(strain_id = group_indices(.)) %>%
  ungroup() %>%group_by(x1_serum, x2_serum) %>%
  mutate(serum_id = group_indices(.))

titer_df_2D$titer = foreach(xi1 = titer_df_2D$x1_strain,
        xi2 = titer_df_2D$x2_strain,
        xk1 = titer_df_2D$x1_serum,
        xk2 = titer_df_2D$x2_serum,
        .combine = 'c') %do% {
          get_one_titer(z_max = 10, xi = c(xi1, xi2), xk = c(xk1, xk2), ndim = 2, c = 1)
        }


titer_mat_2D = matrix(titer_df_2D$titer, nrow = 100, ncol = 100, byrow = F, dimnames = list(paste0('serum', 1:100), paste0('strain', 1:100)))
```

## Perform PCA on 1D titers
```{r}
PCA_2D <- prcomp(x = titer_mat_2D, scale = TRUE, center = TRUE)
summary(PCA_2D)
```

Even in the 1D model, we need 2d to represent the titer map.
Presumably, d1 represents the linear decrease in titer with distance, and d2 represents the peak titer for each dimension?

## Visualize
```{r}
ggbiplot::ggbiplot(PCA_1D, labels = rownames(titer_mat_1D)) +
  ylim(c(-2, 2)) + xlim(c(-2,2))
```