---
title: "What is a dimension"
output: html_notebook
---

## In an antigenic map, what is a dimension?


Immagine a set of 5 strains, labeled a:d.
Immagine that the coordinate positions of these strains represent their antigenic phenotype.
Each unit of Euclidean distance between two strains represents one unit difference in log2 titer.

Now, immagine that strain d mutates to become a new strain, e, which falls 1 distance unit away from d.

**In a 1D map**, a 1-unit differene between d and e means that if the location of d is fixed, there are only two possible locations for e on the map.

The likelihood of each candidate location depends on e's distance from other strains in the panel, a, b, and c.

For example, the following distance matrix would indicate 


```{r}
df_1d <- tibble(strain = c('a', 'b', 'c', 'd'),
                x1 = c(1, 2, 3, 4),
                color_index = as.character(c(1,1,1,1)))
ggplot(df_1d) +
  geom_hline(aes(yintercept = 0))+
  geom_label(aes(x = x1, y = 0, color = color_index, label = strain), show.legend = F) +
  theme(axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank()) +
  ylab('') +
  geom_segment(aes(x = 4, xend = 4.95, y = .01, yend = .01), arrow = arrow(), color = 'cadetblue3') +
  geom_segment(aes(x = 4, xend = 3.05, y = .02, yend = .02), arrow = arrow(), color = 'cadetblue3') +
  geom_text(aes(x = 5, y = .01, label = 'e'), color = 'cadetblue3') +
  geom_text(aes(x = 3, y = .02, label = 'e'), color = 'cadetblue3') +
  ylim(c(0, .05)) +
  ggtitle('1D map:\nTwo possible locations for e')
```



Now, immagine that strain e falls 3 units from d, but only 0.2 units from b. In other words, the mutation dramatically changed the titer to d, but had little effect on the titer to strain b. 

It is not possible to represent this change in a 1D map.

But if we add another dimension, we can represent a mutation that has different effects on the titer to specific strains.

The intesection of two cirles, one centered at d with radius 3, and one centered at b with radius 0.2, represent possible locations for e:


```{r}
df_2d <- tibble(strain = c('a', 'b', 'c', 'd'),
                x1 = c(1, 2, 3, 4),
                x2 = c(3.07, 2.6, 2.75, 0.25),
                color_index = as.character(c(1,1,1,1)))
ggplot(df_2d) +
  geom_label(aes(x = x1, y = x2, color = color_index, label = strain), show.legend = F) +
  ylim(c(-3, 4)) + xlim(c(0,7)) +
  ggforce::geom_circle(aes(x0 = 4, y0 = .25, r = 3), lty = 2) +
  ggforce::geom_circle(aes(x0 = 2, y0 = 2.6, r = .2), lty = 2) +
  geom_point(aes(x = 1.9, y = 2.4), color = 'greenyellow', size = 2) +
  geom_point(aes(x = 2.2, y = 2.6), color = 'greenyellow', size = 2) +
ggtitle('2D map:\nTwo possible locations for e')
```


Following on this logic, if e is 3 units from d, 0.2 units from b, and 1 unit from c, then we need a 3rd dimension to account for the fact that e is close to b, but far from c.

Overall, the extent to which the pairwise distances between strains covary is related to the number of dimensions.





 

