---
title: "Repeated ferret samples"
output: html_notebook
---

## Analyze data from Yoshi K's lab

Experimental design:


# Goals

## 1. Characterize individual variation in measured distances among ferrets infected by the same strain.

There are 3 ferrets per group (treated with the same dose and series of immunization strains).
There are two technical replicated (repeated titer measurements) per ferret.

## 2. Do pairwise distances change between 3 weeks and 2-3 months after immunization?

## 3. Does antiserum's breadth increase between 3 weeks and 2-3 months after immunization?
    
**Background**: This (study of SARS-CoV-2 mABs)[https://www.cell.com/immunity/fulltext/S1074-7613(21)00294-6?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1074761321002946%3Fshowall%3Dtrue] shows that breadth increases from 1.2-6 months post infection. 

In this dataset we can compare antisera taken 3 weeks post infection (consistent with the sampling lag used for most antigenic maps), and approximately 1-2 months later, prior to the next immunization.

We will have limited ability to address question #3, because titers are only measured to three closely related strains.


## Load the data
```{r}
rm(list = ls())
library(readr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(dplyr)

parse_immune_state = function(sample_label, strain_i1_i2, strain_i3, days_since_last_immunization){
  stopifnot(is.factor(sample_label))
  ifelse(sample_label == 'pre immunization 1', 'naive',
         ifelse(sample_label == 'post immunization 1' | sample_label == 'pre immunization 2', sprintf('1x %s +%id', strain_i1_i2, days_since_last_immunization),
                ifelse(sample_label == 'post immunization 2' | sample_label == 'pre immunization 3', sprintf('2x %s +%id', strain_i1_i2, days_since_last_immunization),
                       ifelse(sample_label == 'post immunization 3', sprintf('2x %s\n1x %s +%id', strain_i1_i2, strain_i3, days_since_last_immunization), NA))))
}

get_days_since_last_immunization <- function(sample_date){
  immunization_dates = read_csv('../raw-data/study_design.csv', col_types = 'ccccc') %>%
    dplyr::filter(grepl('immunization', event)) %>%
    mutate(date = as_date(date, format = '%m/%d/%Y')) %>%
    pull(date)
  sapply(sample_date, function(dd){
    immunization_before_sample = ymd(immunization_dates)<ymd(dd)
    if(!any(immunization_before_sample)){
      return(NA)
    }else{
      past_immunization_dates = immunization_dates[immunization_before_sample]
      last_immunization_date = max(past_immunization_dates)
      return(ymd(dd)-ymd(last_immunization_date))
    }
  })
}
```


```{r}
ferret_data = read_csv('../raw-data/machine_readable_raw_data.csv', col_types = 'cccicdidddddddd') %>%
  mutate(date = as_date(date, format = '%m/%d/%Y')) %>% ## cast to date
  pivot_longer(8:15, names_to = c('test_strain', 'replicate'), names_sep = '_', values_to = 'logtiter') %>% ## convert repeated measures of different strains to long format
  mutate(titer = 10*(2^logtiter), ## calculate absolute titer from logtiter
         sample_label = factor(sprintf('%s immunization %i', sample_type, sample_immunization),
                               levels = c('pre immunization 1',  ## Order chronologically
                                          'post immunization 1', 
                                          'pre immunization 2', 
                                          'post immunization 2', 
                                          'pre immunization 3', 
                                          'post immunization 3')),
         last_immunization_strain = ifelse(sample_label == 'pre immunization 1', 'naive', ## Label the last strain with which the ferret was immunized 
                                           ifelse(sample_label == 'post immunization 3', strain_i3, 
                                                  strain_i1_i2)),
         days_since_last_immunization = get_days_since_last_immunization(date),
         immune_state_label = parse_immune_state(sample_label, strain_i1_i2, strain_i3, days_since_last_immunization),
         group = ifelse(strain_i1_i2 == 'Neth399', 'Neth399(WU95 cluster)->Neth312(FU02 cluster)', 'Neth312(FU02 cluster)->Wis15(PE09 cluster)'),
         ferret_id = factor(ferret_id),
         test_strain = factor(test_strain, levels = c('Neth399', 'Neth312', 'Perth09', 'Wis15'))) %>%
  rename(dose = `dose(ug_HA/ferret)`)
```

```{r}
ferret_data %>%
  group_by(sample_label, last_immunization_strain, days_since_last_immunization, immune_state_label) %>%
  summarise()
```



## Variability among technical replicates
```{r}
ferret_data %>%
  group_by(sample_label, ferret_id, test_strain, dose, group) %>%
  summarise(difference = as.factor(diff(logtiter))) %>%
  dplyr::filter(!is.na(difference)) %>%
  ungroup %>%   group_by(sample_label, ferret_id, test_strain, dose, group, difference) %>%
  summarise(n = n()) %>%
  ggplot() +
  #geom_bar(aes(x = test_strain, y = n, color = difference, fill = difference), stat = 'identity') +
  geom_bar(aes(x = difference), breaks = c(-.5, 3.5, by = 1)) +
  facet_grid(dose~test_strain, labeller = label_both) +
  scale_color_viridis_d(aesthetics = c('color', 'fill'))
ggsave('../plots/technical_replicates.png')
```

Among techincal replicates, defined as repeated titer measurements performed on the same sample, measured titers rarely differ by more than a factor of 2.


## Variability among ferrets

```{r}
these_data = ferret_data %>%
  ungroup() %>% group_by(sample_label, test_strain, dose, group) %>%
  mutate(ferret_index = as.factor(as.numeric(factor(as.character(ferret_id))))) %>%
  ungroup() %>%  group_by(sample_label, test_strain, ferret_id) %>%
  mutate(group_id = cur_group_id()) %>% arrange(group_id) %>%
  summarise(log_gmt = mean(logtiter),
            dose = unique(dose),
            group = unique(group),
            ferret_index = unique(ferret_index),
            last_immunization_strain = unique(last_immunization_strain),
            immune_state_label = unique(immune_state_label),
            date = unique(date),
            lag = unique(days_since_last_immunization)) %>%
  ungroup() %>%
  mutate(xval = as.numeric(test_strain)+(as.numeric(ferret_index)-2)*.1,
         last_immunization_strain = factor(last_immunization_strain, levels = c(levels(test_strain), 'naive')),
         dose = factor(dose)) %>%
  group_by(ferret_id, test_strain) %>%
  arrange(sample_label) %>%
  mutate(lag_gmt = dplyr::lag(log_gmt), 
         boost = log_gmt - dplyr::lag(log_gmt)) 

ymax = max(these_data$log_gmt, na.rm = T)

these_data %>%
  ggplot()+
  geom_blank(aes(x = test_strain, y = log_gmt))+
  geom_point(aes(x = xval, y = jitter(log_gmt), color = dose), alpha = .5)+
  geom_line(aes(x = xval, y = log_gmt, color = dose, group = interaction(dose, ferret_index))) +
  geom_vline(aes(xintercept = last_immunization_strain), lty = 2)+
  geom_text(aes(x = 0.5, y = ymax, label = immune_state_label), hjust = 0)+
  ylim(c(0, ymax+1))+
  facet_grid(sample_label~group)
ggsave('../plots/raw_ferret_titers.png', width = 8, height = 10, units = 'in', dpi = 200)
```

```{r}
vline_df = read_csv('../raw-data/study_design.csv', show_col_types = F) %>% 
  dplyr::filter(grepl('immunization', event)) %>%
  select(event, date) %>%
  merge(these_data %>% ungroup() %>% select(group, test_strain)  %>% distinct()) %>%
  mutate(strain2 = gsub(pattern = '.{23}(\\w{5,6}).\\w+ \\w+.', replacement = '\\1', group),
         strain1 = gsub(pattern = '(\\w{7}).+', replacement = '\\1', group),
         vline_date = ifelse(event == 'immunization 3' & as.character(test_strain) == strain2, date,
                               ifelse(event != 'immunization 3' & as.character(test_strain) == strain1, date, NA)),
         vline_date = as_date(vline_date, format = '%m/%d/%Y')) 

these_data %>%
  ggplot()+
  geom_blank(aes(x = date, y = log_gmt))+
  geom_vline(aes(xintercept = vline_date), lty = 2, data = vline_df)+
  geom_point(aes(x = date, y = jitter(log_gmt), color = dose), alpha = .5)+
  geom_line(aes(x = date, y = log_gmt, color = dose, group = interaction(dose, ferret_index))) +
  #geom_text(aes(x = 0.5, y = ymax, label = immune_state_label), hjust = 0)+
  ylim(c(0, ymax+1))+
  facet_grid(test_strain~group)
ggsave('../plots/raw_ferret_titers_chronological.png', width = 8, height = 8, units = 'in', dpi = 200)

```


## Change from prior timepoint among ferrets

```{r}
ymax = max(these_data$boost, na.rm = T)

these_data %>%
  ggplot()+
  geom_blank(aes(x = test_strain, y = boost))+
  geom_point(aes(x = xval, y = jitter(boost), color = dose), alpha = .5)+
  #geom_line(aes(x = xval, y = boost, color = dose, group = interaction(dose, ferret_index))) +
  geom_vline(aes(xintercept = last_immunization_strain), lty = 2)+
  geom_text(aes(x = 0.5, y = ymax, label = immune_state_label), hjust = 0)+
  ylim(c(0, ymax+1))+
  facet_grid(sample_label~group)
ggsave('../plots/raw_ferret_titers_boost.png', width = 8, height = 10, units = 'in', dpi = 200)
```


## Changes from samples 2-3, and 4-5
```{r}
these_data %>%
  mutate(xval = (as.numeric(sample_label)+((as.numeric(dose)-2)*.15)+((as.numeric(ferret_index)-2)*.02))) %>%
  dplyr::filter(sample_label %in% c('post immunization 1', 'pre immunization 2')) %>%
  group_by(ferret_index, dose, test_strain, group) %>%
  arrange(sample_label) %>%
  mutate(norm_gmt = log_gmt - log_gmt[1]) %>%
  dplyr::filter(!grepl('Wis', test_strain)) %>%
  ggplot() +
  geom_hline(aes(yintercept = 0), lty = 2)+
  geom_point(aes(x = xval, y = norm_gmt, color = dose))+
  geom_line(aes(x = xval, y = norm_gmt, color = dose, group = interaction(ferret_index, dose)), alpha = .5)+
  geom_label(aes(x = 2, y = 1.5, label = sprintf('+%sd', lag)))+
  facet_grid(test_strain~group) +
  ylab('change in gmt') +
  xlab('sampling timepoint') +
  ylim(c(-2, 2)) +
  scale_x_continuous(breaks=c(2, 3),
        labels=c("post immunization 1", "pre immunization 2"))+
  theme(axis.text.x = element_text(angle = 10))
ggsave('../plots/compare_poast_i1.png', width = 8, height = 8, units = 'in', dpi = 200)
```

```{r}
these_data %>%
  mutate(xval = (as.numeric(sample_label)+((as.numeric(dose)-2)*.15)+((as.numeric(ferret_index)-2)*.02))) %>%
  dplyr::filter(sample_label %in% c('post immunization 2', 'pre immunization 3')) %>%
  group_by(ferret_index, dose, test_strain, group) %>%
  arrange(sample_label) %>%
  mutate(norm_gmt = log_gmt - log_gmt[1]) %>%
  ggplot() +
  geom_hline(aes(yintercept = 0), lty = 2)+
  geom_point(aes(x = xval, y = norm_gmt, color = dose))+
  geom_line(aes(x = xval, y = norm_gmt, color = dose, group = interaction(ferret_index, dose)), alpha = .5)+
  geom_label(aes(x = 4, y = 1.5, label = sprintf('+%sd', lag)))+
  facet_grid(test_strain~group) +
  ylab('change in gmt') +
  xlab('sampling timepoint') +
  ylim(c(-2, 2)) +
  scale_x_continuous(breaks=c(4,5),
        labels=c("post immunization 2", "pre immunization 3"))+
  theme(axis.text.x = element_text(angle = 10))
ggsave('../plots/compare_poast_i2.png', width = 8, height = 10, units = 'in', dpi = 200)
```



```{r}
these_data %>%
  mutate(is_homologous_strain = as.character(test_strain) == as.character(last_immunization_strain),
         temporal_group = gsub(pattern = '(.{2}).+', '\\1', immune_state_label)) %>%
  dplyr::filter(sample_label %in% c('post immunization 1', 'pre immunization 2', 'post immunization 2', 'pre immunization 3')) %>%
  group_by(temporal_group, ferret_id, test_strain, is_homologous_strain) %>%
  arrange(date) %>%
  summarize(log_titer_change = log_gmt[2]-log_gmt[1],
            dt = date[2]-date[1]) %>%
  ggplot() +
  geom_hline(aes(yintercept = 0), lty = 2)+
  geom_hline(aes(yintercept = 1), lty = 3)+
  geom_hline(aes(yintercept = -1), lty = 3)+
  geom_violin(aes(x = temporal_group, y = log_titer_change, color = is_homologous_strain), draw_quantiles = c(.25, .5, .75)) +
  ylim(c(-2, 2))
```

