---
title: "Muti epitope maps -- ferrets"
output: html_notebook
---



**Workflow:**

1. Simulate a three-epitope pathogen with 5 antigens of each epitope represented on a 2D map.
2. Simulate ferret antibody repertoires to the epitope.
3. Calculate each ferret's titer to each strain to simulate a sero panel.

------

Next script:

4. From the sero panel, infer the "strain map" in 1, 2, 3, ... 8 dimensions.
5. Repeat under conditions of even, skewed, and variable immunodominance.
6. Calculate the map error for each fit.



```{r}
source('../../R/utility-functions.R')
source('../../R/multi_epitope_code_sparse_dataset.R')
library('foreach')



n_antigens = 5
n_ab = 0
n_chains = 4
n_iter = 5000 # half will be burn in
n_epitopes = 3

```

## Generate antigen coordinates. Each antigen consists of 3 epitopes.
```{r}
set.seed(13)
antigen_coords <- tibble(epitope = rep(c(1:n_epitopes), each = n_antigens),
                         antigen = rep(1:n_antigens, n_epitopes),
                         kind = 'antigen',
                         c1 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c2 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c3 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c4 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c5 = generate_ag_coords_random(n_antigens, n_epitopes)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))


## Shift so that epitope 1 of antigen 1 falls at the origin
shift_this_col = function(xx){ xx - xx[1] }
antigen_coords <- antigen_coords %>%
  mutate_at(.vars = vars(matches('c\\d?.+')), .funs = shift_this_col)
## Set all coords of ag2 epitope 1 to 0 other than c1, so that the ag falls on the x axis
modify_these_columns = grep(names(antigen_coords), pattern = 'c\\d?.+')[-1]
antigen_coords[2, modify_these_columns] = 0

write_csv(antigen_coords[,1:5], paste0('antigen_coords.csv'))
write_csv(antigen_coords[,1:6], paste0('antigen_coords_3D.csv'))
write_csv(antigen_coords[,1:7], paste0('antigen_coords_4D.csv'))
write_csv(antigen_coords[,1:8], paste0('antigen_coords_5D.csv'))
```

## Plot the coordinates
```{r}
antigen_coords[,1:5] %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), alpha = .2)
ggsave(filename = paste0('epitope_coords_polygon.png'))
```

```{r}
antigen_coords[,1:5] %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  facet_wrap(.~epitope)
ggsave(filename = paste0('epitope_coords_grid.png'))
```
## Generate sera and titer inputs
```{r}
## Write a function to generate and save one set of inputs for each number of dimensions per epitope (n_dim) and immunodomiannce scheme
generate_inputs_wrapper <- function(n_dim,
                                    immunodominance_flag,
                                    immunodominance_vec){
  generate_ferret_inputs(antigen_coords, 
                         relative_immunodominance = immunodominance_vec, 
                         n_epitopes = 3, 
                         n_antigens = 5, 
                         n_dim = n_dim, 
                         immunodominance_flag = immunodominance_flag, 
                         outdir = 'inputs',
                         n_abs_per_serum = 1000, 
                         sigma = .1)
}
```

## Generate inputs for each immunodominance scheme and dimensions
```{r}
input_generation_list <- expand_grid(n_dim = 2:5,
                                     immunodominance_flag = c('even', 'skewed', 'E1', 'E2', 'E3'),
                                     immunodominance_vec = list(c(1,1,1),
                                                                c(1+2/3, 1, 1/3),
                                                                c(3,0,0),
                                                                c(0,3,0),
                                                                c(0,0,3)))

foreach(this_n_dim = input_generation_list$n_dim,
        this_immunodominance_flag = input_generation_list$immunodominance_flag,
        this_immunodominance_vec = input_generation_list$immunodominance_vec) %do% {
          generate_inputs_wrapper(n_dim = this_n_dim, 
                                  immunodominance_flag = this_immunodominance_flag, 
                                  immunodominance_vec = this_immunodominance_vec)
        }
```