---
title: "Muti epitope maps -- ferrets"
output: html_notebook
---



**Workflow:**

1. Simulate a three-epitope pathogen with 5 antigens of each epitope represented on a 2D map.
2. Simulate ferret antibody repertoires to the epitope.
3. Calculate each ferret's titer to each strain to simulate a sero panel.

------

Next script:

4. From the sero panel, infer the "strain map" in 1, 2, 3, ... 8 dimensions.
5. Repeat under conditions of even, skewed, and variable immunodominance.
6. Calculate the map error for each fit.



```{r}
source('../../R/utility-functions.R')
source('../../R/multi_epitope_code_sparse_dataset.R')
library('foreach')



n_antigens = 5
n_ab = 0
n_chains = 4
n_iter = 5000 # half will be burn in
n_epitopes = 3

```

## Generate antigen coordinates. Each antigen consists of 3 epitopes.
```{r}
set.seed(13)
antigen_coords <- tibble(epitope = rep(c(1:n_epitopes), each = n_antigens),
                         antigen = rep(1:n_antigens, n_epitopes),
                         kind = 'antigen',
                         c1 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c2 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c3 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c4 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c5 = generate_ag_coords_random(n_antigens, n_epitopes)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))


## Shift so that epitope 1 of antigen 1 falls at the origin
shift_this_col = function(xx){ xx - xx[1] }
antigen_coords <- antigen_coords %>%
  mutate_at(.vars = vars(matches('c\\d?.+')), .funs = shift_this_col)
## Set all coords of ag2 epitope 1 to 0 other than c1, so that the ag falls on the x axis
modify_these_columns = grep(names(antigen_coords), pattern = 'c\\d?.+')[-1]
antigen_coords[2, modify_these_columns] = 0

write_csv(antigen_coords[,1:5], paste0('antigen_coords.csv'))
write_csv(antigen_coords[,1:6], paste0('antigen_coords_3D.csv'))
write_csv(antigen_coords[,1:7], paste0('antigen_coords_4D.csv'))
write_csv(antigen_coords[,1:8], paste0('antigen_coords_5D.csv'))
```

## Plot the coordinates
```{r}
antigen_coords[,1:5] %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), alpha = .2)
ggsave(filename = paste0('epitope_coords_polygon.png'))
```

```{r}
antigen_coords[,1:5] %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  facet_wrap(.~epitope)
ggsave(filename = paste0('epitope_coords_grid.png'))
```
## Generate sera and titer inputs
```{r}
even_immunodominance_inputs <- generate_ferret_inputs(antigen_coords[,1:5], 
                                                      relative_immunodominance = c(1, 1, 1), 
                                                      n_epitopes = 3, 
                                                      n_antigens = 5, 
                                                      n_abs_per_serum = 1000, 
                                                      sigma = .1)
```


```{r}
skewed_immunodominance_inputs <- generate_ferret_inputs(antigen_coords[,1:5], 
                                                      relative_immunodominance = c(1+2/3, 1, 1/3), 
                                                      n_epitopes = 3, 
                                                      n_antigens = 5, 
                                                      n_abs_per_serum = 1000, 
                                                      sigma = .1)
```


```{r}
E1_complete_dominance_inputs <- generate_ferret_inputs(antigen_coords[,1:5], 
                                                      relative_immunodominance = c(3, 0, 0), 
                                                      n_epitopes = 3, 
                                                      n_antigens = 5, 
                                                      n_abs_per_serum = 1000, 
                                                      sigma = .1)
```

```{r}
E2_complete_dominance_inputs <- generate_ferret_inputs(antigen_coords[,1:5], 
                                                      relative_immunodominance = c(0, 3, 0), 
                                                      n_epitopes = 3, 
                                                      n_antigens = 5, 
                                                      n_abs_per_serum = 1000, 
                                                      sigma = .1)
```

```{r}
E3_complete_dominance_inputs <- generate_ferret_inputs(antigen_coords[,1:5], 
                                                      relative_immunodominance = c(0, 0, 3), 
                                                      n_epitopes = 3, 
                                                      n_antigens = 5, 
                                                      n_abs_per_serum = 1000, 
                                                      sigma = .1)
```

## Save inputs
```{r}
write_rds(even_immunodominance_inputs, paste0(outdir, '/even_immunodominance_inputs.rds'))
write_rds(skewed_immunodominance_inputs, paste0(outdir, '/skewed_immunodominance_inputs.rds'))
write_rds(E1_complete_dominance_inputs, paste0(outdir, '/E1_complete_dominance_inputs.rds'))
write_rds(E2_complete_dominance_inputs, paste0(outdir, '/E2_complete_dominance_inputs.rds'))
write_rds(E3_complete_dominance_inputs, paste0(outdir, '/E3_complete_dominance_inputs.rds'))
```
