---
title: "Bayesian MDS in Stan"
output: html_notebook
---

Write and test a Stan model

```{r}
generate_alleles <- function(ndim, nalleles, range = 10){
  # Randomly draw coordinates in each dimension
  sapply(1:ndim, function(dim){runif(nalleles, -range/2, range)}) %>%
    as.tibble() %>%
    set_names(paste0('c', 1:ndim)) %>%
    # Assign allele ids
    mutate(allele = 1:nalleles)
}
```

```{r}
## Generate coordinates for a separate set of antigens (Ag) and sera (Ab)
ab_df <- generate_alleles(2, 5)
ag_df <- generate_alleles(2, 5)

ab_coords <- as.matrix(ab_df %>% select(starts_with('c')))
ag_coords <- as.matrix(ag_df %>% select(starts_with('c')))

ag_ab_df<- bind_rows(list('ag' = ag_df,
               'ab' = ab_df),
          .id = 'kind') %>%
  rename(id = allele) %>%
  select(kind, id, starts_with('c'))

true_plot <- ag_ab_df %>%
  ggplot(aes(x = c1, y = c2, color = kind)) +
  geom_text(aes(label = id))

true_plot
```

```{r}
get_l2_distance<- function(v1, v2){
  stopifnot(length(v1) == length(v2))
  sqrt(sum((v1-v2)^2))
}

## Calculate the distance matrix
get_distances <- function(ab_coords, ag_coords){
  distmat = matrix(NA, nrow = nrow(ag_coords), ncol = nrow(ab_coords))
  rownames(distmat) = paste0('ag', 1:nrow(ag_coords))
  colnames(distmat) = paste0('ab', 1:nrow(ab_coords))
  for(ab in 1:nrow(ab_coords)){
    for(ag in 1:nrow(ag_coords)){
      distmat[ag, ab] = get_l2_distance(ab_coords[ab,], ag_coords[ag,])
    }
  }
  distmat
}
```

```{r}
observed_distances = get_distances(ab_coords, ag_coords)

## Test
# sqrt(sum((ab_coords[3,]-ag_coords[5,])^2)) == observed_distances[5,3]
# sqrt(sum((ab_coords[1,]-ag_coords[1,])^2)) == observed_distances[1,1]
# sqrt(sum((ab_coords[2,]-ag_coords[5,])^2)) == observed_distances[5,2]
# sqrt(sum((ab_coords[2,]-ag_coords[4,])^2)) == observed_distances[4,2]
# sqrt(sum((ab_coords[2,]-ag_coords[3,])^2)) == observed_distances[3,2]
```


```{r}
fit_MDS_MLE <- function(
  observed_distances, # vector of delay for each individual in the data
  n_antigens,
  n_sera,
  n_dim
) {

  stopifnot(nrow(observed_distances) == n_antigens)
  stopifnot(ncol(observed_distances) == n_sera)
  
  loss_function <- function(pars, observed_distances, n_antigens, n_sera, n_dim){
    stopifnot(length(pars) == (n_antigens+n_sera)*n_dim)
    
    estimated_ag_coords <- matrix(pars[1:(n_antigens*n_dim)], n_antigens, n_dim)
    estimated_ab_coords <- matrix(pars[(n_antigens*n_dim+1):length(pars)], n_sera, n_dim)
    
    estimated_distances <- get_distances(estimated_ab_coords, estimated_ag_coords)
    sum((observed_distances - estimated_distances)^2)
  }

  generate_initial_guess <- function(n_antigens, n_sera, n_dim){
    ag_c1s <- runif(n_antigens, -10, 10); names(ag_c1s) = paste0('ag', 1:n_antigens, 'c1')
    ag_c2s <- runif(n_antigens, -10, 10); names(ag_c2s) = paste0('ag', 1:n_antigens, 'c2')
    ab_c1s <- runif(n_sera, -10, 10); names(ab_c1s) = paste0('ab', 1:n_antigens, 'c1')
    ab_c2s <- runif(n_sera, -10, 10); names(ab_c2s) = paste0('ab', 1:n_antigens, 'c2')
    
    c(ag_c1s, ag_c2s, ab_c1s, ab_c2s)
  }
  
  ## Repeat the fit 5 times from different initial values to verify convergence
  
  fit_list <- lapply(1:10, function(x) { optim(par = generate_initial_guess(n_antigens, n_sera, n_dim),
                                             fn = loss_function,
                                             observed_distances = observed_distances,
                                             n_antigens = n_antigens, 
                                             n_sera = n_sera,
                                             n_dim = n_dim,
                                             method = 'CG',
                                             control = list(maxit = 500))}
  )
  ## Check convergence
  if(!all(sapply(fit_list, function(ll) ll$convergence) == 0)){
    warning('Not all optimizations converged! - Check individual fits')
    return(fit_list)
  }
  ## Check that if all runs converged, that they all converged to the same solution
  errfun_value <- sapply(fit_list, function(ll) ll$value)
  if(!all(abs(errfun_value - mean(errfun_value) < .5))){warning('Not all runs converged to the same solution. Check individual fits.')}
  
  return(list(best_fit = fit_list[errfun_value == min(errfun_value)][[1]],
              all_fits = fit_list))
}

```

```{r}
fits <- fit_MDS_MLE(observed_distances = observed_distances, n_antigens = 5, n_sera = 5, n_dim = 2)
fits
```
## Visualize the difference between fits
```{r}
get_fits_df <- function(fit_obj){
  tibble(coordinate = fit_obj$par,
       label = names(fit_obj$par)) %>%
  tidyr::extract(label, into = c('kind','id', 'dim'), regex = '(a\\w)(\\d)c(\\d)', convert = T) %>%
  pivot_wider(names_from = dim, values_from = coordinate, names_prefix = 'c') 
}
```


Plot the solutions from each run.

Note that the runs with near-0 errfun all have very similar topology, whereas the higher-errfun solutions seem like a local min solution
```{r}
all_fits_df <- lapply(fits$all_fits, function(ll){ 
  get_fits_df(ll) %>%
  mutate(errfun = ll$value,
         convergence = ll$convergence) }) %>%
  bind_rows(.id = 'fit')
```

```{r}
all_fits_df %>%
  ggplot()+
  geom_point(aes(x = c1, y = c2, shape = kind, color = errfun)) +
  facet_wrap(.~fit)
```


## Compare errorfunctions
```{r}
tibble(fit = 1:10,
       value = sapply(fits$all_fits, function(ll) ll$value)
) %>%
  ggplot() +
  geom_histogram(aes(x = value), bins = 100)
```



## Compare distance matrices
```{r}
get_distances(ab_coords = fits_df %>% dplyr::filter(kind == 'ab') %>% select(starts_with('c')) %>% as.matrix,
              ag_coords = fits_df %>% dplyr::filter(kind == 'ag') %>% select(starts_with('c')) %>% as.matrix)

observed_distances
```

## Plot
```{r}
fits_df <- get_fits_df(fits$best_fit)

mds_plot <- fits_df %>%
  ggplot(aes(x = c1, y = c2, color = kind))+
  geom_text(aes(label = id))

true_plot %>% ggtitle('truth')
```
```{r}
mds_plot %>% ggtitle('mds')
```

## The MDS can recover the map, but it's often shifted, rotated, and reflected relative to the original

## Try fixing one point to see if we can recover the original better.



NOTE - below I tried to change basis to recover the original map, but the basis transformation is distorting the MDS solution and this isn't working.



## Transform to the standard basis:

```{r}
## Solve for the basis transform
solve_for_transition_to_standard_basis <- function(mds_coords, original_coords){
  loss_function <- function(pars, mds_coords, original_coords){
    v1 = pars['v1']
    v2 = pars['v2']
    u1 = pars['u1']
    u2 = pars['u2']
    
    Cprime = t(mds_coords)
    C = t(original_coords)
    
    transition_matrix = matrix(c(v1, v2, u1, u2), 2, 2)
    
    sqrt(sum(((transition_matrix %*% Cprime)-C)^2))
  }
  
  optim(par = c(v1 = runif(1, -5, 5), v2 = runif(1, -5, 5), u1 = runif(1, -5, 5), u2 = runif(1, -5, 5)), fn = loss_function, mds_coords = mds_coords, original_coords = original_coords)
}
```

```{r}
## Transform to standard basis
transform_to_standard_basis <- function(mds_coords, original_coords){
  stopifnot(all(dim(mds_coords) == dim(original_coords)))
  transition_coords <- solve_for_transition_to_standard_basis(mds_coords, original_coords)
  transition_matrix <- matrix(transition_coords$par, 2, 2)
  transition_matrix %*% t(mds_coords) %>%
    t() %>%
    as.tibble() %>%
    # Vertical and horizontal shift to re-center
    mutate(V1 = V1 - mean(V1) + mean(original_coords[,1]),
           V2 = V2 - mean(V2) + mean(original_coords[,2])) %>%
    set_names(paste0('c', 1:2)) %>%
    mutate(id = 1:n())
}
```

```{r}
## Apply function
transformed_mds <- transform_to_standard_basis(mds_coords = fits_df %>% select(starts_with('c')) %>% as.matrix(), 
                            original_coords = rbind(ag_coords, ab_coords)) %>%
  mutate(kind = rep(c('ag', 'ab'), each = 5),
         id = ag_ab_df$id) %>%
  select(kind, id, starts_with('c'))
```


```{r}
get_distances(ab_coords = transformed_mds %>% dplyr::filter(kind == 'ab') %>% select(starts_with('c')) %>% as.matrix,
              ag_coords = transformed_mds %>% dplyr::filter(kind == 'ag') %>% select(starts_with('c')) %>% as.matrix)
get_distances(ab_coords = fits_df %>% dplyr::filter(kind == 'ab') %>% select(starts_with('c')) %>% as.matrix,
              ag_coords = fits_df %>% dplyr::filter(kind == 'ag') %>% select(starts_with('c')) %>% as.matrix)
observed_distances
```




## Visualize fits using mean estimate for each coord
```{r}
bind_rows(list('raw_mds' = fits_df, 
              'actual' = ag_ab_df,
              'transformed_mds' = transformed_mds
          ),
          .id = 'source') %>%
  mutate(panel = ifelse(source == 'raw_mds', 2, 1)) %>%
  ggplot() +
  geom_text(aes(x = c1, y = c2, color = source, label = id)) +
  facet_grid(.~panel)
```


--> This isn't really working. I'm not sure if it's a problem with the model or a mean/median posterior thing.
