---
title: "v0.2_test_MLE"
author: "Katie Gostic"
date: "1/20/2022"
output: html_document
---

## Test the least squares MLE approach
* use functions developed in v0.1
* but this time don't fix the first Ag coordinate

```{r setup, include=FALSE}
rm(list = ls())
library(tidyr)
filter <- dplyr::filter
extract <- tidyr::extract
source('../code.R')
```

## Generate synthetic coordinates
```{r cars}
coords <- generate_map_coordinates(n_dim = 2, n_antigens = 5, n_sera = 6, map_range = c(-10, 10))
```

## Plot
```{r}
coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = kind))
```
## Get distances
```{r}
distmat <- get_ab_ag_distances(ab_coords = filter(coords, kind == 'serum') %>% coord_tibble_to_matrix(),
                               ag_coords = filter(coords, kind == 'antigen') %>% coord_tibble_to_matrix())
```

## Use MSD to infer coordinates from distances
```{r}
fit_list <- fit_MDS_least_squares(observed_distances = distmat, n_antigens = nrow(distmat), n_sera = ncol(distmat), n_dim = 2)
```

## Visualize the 10 fits, each from different initial conditions
```{r}
get_fits_df <- function(fit_obj){
  tibble(coordinate = fit_obj$par,
         label = names(fit_obj$par)) %>%
    tidyr::extract(label, into = c('kind','id', 'dim'), regex = '(a\\w)(\\d)c(\\d)', convert = T) %>%
    pivot_wider(names_from = dim, values_from = coordinate, names_prefix = 'c') 
}


all_fits_df <- lapply(fit_list$all_fits, function(ll){ 
  get_fits_df(ll) %>%
  mutate(errfun = ll$value,
         convergence = ll$convergence) }) %>%
  bind_rows(.id = 'fit')
```

Note that not all the fits have the same errfun value. Some get stuck at a local minimum that is not nearly as good as the global minmum.

```{r}
all_fits_df %>%
  ggplot()+
  geom_point(aes(x = c1, y = c2, shape = kind, color = errfun)) +
  facet_wrap(.~fit)
```
## Visualize the best solution next to the original coords

The solution will be an isometry of the original coords
```{r}
best_fit <- get_fits_df(fit_list$best_fit)

bind_rows(list(best_fit = best_fit,
               original = coords),
          .id = 'type') %>%
  ggplot() +
  geom_text(aes(x = c1, y = c2, color = type, label = id))
```

## Translate the mds solution to align with the original coords
```{r}
aligned_mds_coords <- align_mds_with_original(mds_coords = best_fit %>% select(c1, c2) %>% as.matrix(), 
                                              original_coords = coords %>% select(c1, c2) %>% as.matrix())
```

## Visualize alignment between mds coords and original
```{r}
bind_rows(list(original = coords,
               realigned_mds = coords %>% mutate(c1 = aligned_mds_coords[,1], c2 = aligned_mds_coords[,2])),
          .id = 'type') %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, shape = type, color = type)) +
  scale_shape_manual(values = c(16, 3))
```



