---
title: "Muti epitope maps -- combine different immunodominance schemes"
output: html_notebook
---



**Workflow:**

1. Simulate a three-epitope pathogen with 5 antigens of each epitope represented on a 2D map.
2. Simulate ferret antibody repertoires to the epitope, assuming different immunodominance schemes.
3. Calculate each ferret's titer to each strain to simulate a sero panel, combining sera from ferrets with different immunodominance heierarchies.

------


```{r}
rm(list = ls())
source('../../R/multi_epitope_code_sparse_dataset.R')
source('../../R/code.R')
library('foreach')



n_antigens = 5
n_ab = 0
n_chains = 4
n_iter = 5000 # half will be burn in
n_epitopes = 3

outdir = 'simulated_inputs'
if(!dir.exists(outdir)) dir.create(outdir)
```

## Generate antigen coordinates. Each antigen consists of 3 epitopes.
```{r}
set.seed(13)
antigen_coords <- tibble(epitope = rep(c(1:n_epitopes), each = n_antigens),
                         antigen = rep(1:n_antigens, n_epitopes),
                         kind = 'antigen',
                         c1 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c2 = generate_ag_coords_random(n_antigens, n_epitopes)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))


## Standardize so that Ag1 epitope 1 is at the origin and Ag2 epitope 1 is on the x-axis
shifted_coords <-  standardize_coordinates(coord_mat = antigen_coords %>% select(starts_with('c')) %>% as.matrix(), ag1_row = 1, ag2_row = 2) %>%
  as_tibble() %>%
  set_names(c('c1', 'c2'))

antigen_coords$c1 = shifted_coords$c1
antigen_coords$c2 = shifted_coords$c2

write_csv(antigen_coords, paste0(outdir, '/antigen_coords.csv'))
```

## Plot the coordinates
```{r}
antigen_coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), alpha = .2)
ggsave(filename = paste0(outdir, '/epitope_coords_polygon.png'))
```

```{r}
antigen_coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  facet_wrap(.~epitope)
ggsave(filename = paste0(outdir, '/epitope_coords_grid.png'))
```

```{r}
## Generate a data frame with serum metadata - which strain are Abs specific to and which epitope is immunodominant?
serum_catalog <- expand_grid(infecting_strain = 1:n_antigens,
                        dominant_epitope = 1:2,
                        immunodominance_scheme = c('all-or-nothing', 'skewed', 'even'),
                        replicate = 1:3# Do 10 replicates for each possible comination
                        ) %>%
  mutate(serum_id = 1:nrow(.))
```


```{r include=FALSE}
## Generate Ab populaitons for each serum.

serum_list <- foreach(this.infecting.strain = serum_catalog$infecting_strain,
        this.dominant.epitope = serum_catalog$dominant_epitope,
        this.immunodominance = serum_catalog$immunodominance_scheme,
        serum.id = serum_catalog$serum_id) %do% {
          
          #cat(sprintf('serum id is %s\n', serum.id))
          n_epitopes = nrow(unique(antigen_coords[,'epitope']))
          if(this.immunodominance == 'even'){
            immunodominance_vec = rep(1/n_epitopes, n_epitopes)
          }else if(this.immunodominance == 'skewed'){
            immunodominance_vec = dexp(1:n_epitopes, 1)
            immunodominance_vec = immunodominance_vec/sum(immunodominance_vec)
            
          }else{
            stopifnot(this.immunodominance == 'all-or-nothing')
            immunodominance_vec = vector(mode = 'numeric', length = n_epitopes)
            immunodominance_vec[this.dominant.epitope] = 1
          }
          this_epitope_coords <- antigen_coords %>% 
            filter(antigen==this.infecting.strain)
          stopifnot(nrow(this_epitope_coords) == n_epitopes)
          generate_one_serum(native_epitope_coords = this_epitope_coords,
                                   n_epitopes = n_epitopes,
                                   total_ab = 500,
                                   epitope_immunodominance = immunodominance_vec,
                                   strain_dominance = 1, # We assume there's only one strain in the infection history
                                   sigma = .1) %>%
            mutate(serum_id = serum.id)
        } 
```

## Generate a titer map 

* Calculate the titer of each serum to each antigen.

```{r}
titer_map <- matrix(NA, nrow = length(serum_list), ncol = n_antigens, dimnames = list(paste0('serum', serum_catalog$serum_id), paste0('antigen', 1:n_antigens)))
this.row = 0

foreach(serum = serum_list,
        .combine = 'c') %do%{
  this.row = this.row+1
  # Extract serum Ab coords as a list of matrices, one for each eptiope
  this_ab_position_list = lapply(1:n_epitopes, function(this.epitope){
    serum %>% 
      filter(epitope == this.epitope) %>%
      select(matches('c\\d_Ab')) %>% 
      as.matrix()
  })
for(this.antigen in 1:n_antigens){
    # Extract antigen coords as a list of matrices, one for each epitope
    this_antigen_list <- lapply(1:n_epitopes, function(this.epitope){
      antigen_coords %>% 
        filter(epitope == this.epitope & antigen == this.antigen) %>%
        select(starts_with('c')) %>%
        as.matrix()
    })
    # solve for titer  
    #cat(sprintf('This serum is %s, this antigen is %s', this.row, this.antigen))
    titer_map[this.row, this.antigen] = solve_for_titer_multi_epitope(ab_position_list = this_ab_position_list, 
                                                                       relative_concentrations = c(1,1,1), 
                                                                       ag_list = this_antigen_list, 
                                                                       alpha = .25, 
                                                                       r = 7)
  }
}

titer_map
```
## Convert into a long data frame
```{r}
titer_map <- titer_map %>% 
  as.tibble() %>%
  bind_cols(serum_catalog) %>%
  pivot_longer(contains('antigen'), names_to = 'antigen', names_prefix = 'antigen', values_to = 'titer') %>%
  dplyr::rename(serum = serum_id) %>%
  mutate(antigen = as.integer(antigen),
         logtiter = log2(titer/10)) %>% 
    group_by(serum) %>%
    mutate(serum_potency = max(logtiter)) %>%
    ungroup() %>% group_by(antigen) %>%
    mutate(antigen_avidity = max(logtiter)) %>%
    ungroup() %>%
    mutate(titer_distance = (serum_potency+antigen_avidity)/2 - logtiter)
  
```



## Save inputs
```{r}
write_rds(titer_map, paste0(outdir, '/titer_inputs.rds'))
write_rds(serum_catalog, paste0(outdir, '/serum_catalog.rds'))
write_rds(serum_list, paste0(outdir, '/serum_list.rds'))
```
