---
title: "Develop coordinate standardization function"
output: html_notebook
---

Write a function that will take in an arbitrary matrix of 2d or 3d coordinates and standardize them so that:

* Ag1 is at the origin
* Ag2 is on the x-axis

```{r setup, include=FALSE}
rm(list = ls())
library(tidyr)
filter <- dplyr::filter
extract <- tidyr::extract
select <- dplyr::select
source('../code.R')
```


```{r cars}
n_antigens = 5
n_sera = 6
n_dim = 2
coords <- generate_map_coordinates(n_dim = n_dim, 
                                   n_antigens = n_antigens, 
                                   n_sera = n_sera, 
                                   map_range = c(-10, 10))
coords

coord_mat <- coord_tibble_to_matrix(coords)
```

## Plot the synthetic coordinates
```{r}
original_coord_plot <- coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = kind))
```

## Get the distances between coordinates

* We only observe the distances between antigen (ag) and antibody/serum (ab), not between ag-ag or ab-ab.
* Assume all ag-ab distances are observed without error

```{r}
# Like most of the functions in thei script, `get_ab_ag_distances` is a function defined in `../code.R`
distmat <- get_ab_ag_distances(ab_coords = filter(coords, kind == 'serum') %>% coord_tibble_to_matrix(),
                               ag_coords = filter(coords, kind == 'antigen') %>% coord_tibble_to_matrix())
```

## Transform the coordinates so that ag1 is at the origin, and ag2 falls on axis 1:
```{r}
standardize_coordinates <- function(coord_mat, 
                                    ag1_row = 1,
                                    ag2_row = 2
){
  n_dim = ncol(coord_mat) 
  stopifnot(nrow(coord_mat) >= n_dim)
  
  ## Shift the whole matrix so that ag1 is at the origin.
  shifted = shift_to_origin(coord_mat, origin_row = ag1_row)
  # cat('shifted cords\n')
  # print(shifted)
  
  if(n_dim == 3){
  ## The coordinates of the vector (ag1, ag2) are now the ag2 coordinates, because all ag1 coordinates are 0.
  ## We want to rotate the map so that ag2 falls on axis 1
  x_z_projection_vec = c(shifted[2, 1], 0, shifted[2,3])
  theta = get_theta(v1 = c(1,0,0), v2 = x_z_projection_vec)
  theta = ifelse(shifted[2,3]>0, 2*pi-theta, theta) ## Rotate in the opposite direction if z is positive
  rotated_matrix = rotate_about_y_axis_3d(theta = theta, coords = shifted)
  
  ## Now the ag2 vector is in the x-y plane.
  ## Rotate again about the z-axis to align ag2 with the x axis, and return
  theta2 = get_theta(c(rotated_matrix[2,1],0,0), rotated_matrix[2,])
  final_output = rotate_about_z_axis_3d(theta = theta2, 
                                        coords = rotated_matrix)

  # cat('\nafter rotation 1:\n')
  # print(rotated_matrix[2,])
  # cat('\nafter rotation 2:\n')
  stopifnot(equal_ish(final_output[2,2], 0) & equal_ish(final_output[2,3],0))
    
  }else if(n_dim == 2){
    ## Rotate about the origin to align with x and return
    theta = get_theta(shifted[ag2_row,], c(1,0))
    theta = ifelse(shifted[ag2_row,2]>0, 2*pi-theta, theta)
    # cat('theta:\n')
    # print(theta)
    # cat('original:\n')
    # print(shifted[ag2_row,])
    final_output = rotate_by_theta_2d(theta = theta, 
                       coords = shifted)
    stopifnot(equal_ish(final_output[2,2],0))
    
  }else{
    stop('standardization only implemented for 2d or 3d maps')
  }
  return(final_output)
}
```

## Test 2D
```{r}
## Generate a random 2d set of coords and test that they are shifted with distances preserved
test_mat <- generate_map_coordinates(n_dim = 2, 
                                   n_antigens = n_antigens, 
                                   n_sera = n_sera,                              
                                   map_range = c(-10, 10))%>%
 coord_tibble_to_matrix()
original_distances <- get_ab_ag_distances(test_mat[1:n_antigens,], test_mat[(n_antigens+1):(n_antigens+n_sera),])
standardized <- standardize_coordinates(coord_mat = test_mat, 1, 2)
standardized_distances <- get_ab_ag_distances(standardized[1:n_antigens,], standardized[(n_antigens+1):(n_antigens+n_sera),])
## Checks
cat('are distances preserved?\n')
equal_ish(original_distances, standardized_distances)
cat('are coordinates shifted properly?\n')
standardized[1:2,]
```

## Test 3D
```{r}
## Generate a random 3d set of coords and test that they are shifted with distances preserved
test_mat <- generate_map_coordinates(n_dim = 3, 
                                   n_antigens = n_antigens, 
                                   n_sera = n_sera,                              
                                   map_range = c(-10, 10))%>%
 coord_tibble_to_matrix()
original_distances <- get_ab_ag_distances(test_mat[1:n_antigens,], test_mat[(n_antigens+1):(n_antigens+n_sera),])
standardized <- standardize_coordinates(coord_mat = test_mat, 1, 2)
standardized_distances <- get_ab_ag_distances(standardized[1:n_antigens,], standardized[(n_antigens+1):(n_antigens+n_sera),])
## Checks
cat('are distances preserved?\n')
equal_ish(original_distances, standardized_distances)
cat('are coordinates shifted properly?\n')
standardized[1:2,]
```