---
title: "Bayesian MDS in Stan"
output: html_notebook
---


Test basic Bayesian MDS, initially using perfectly observed, idealized data.


```{r}
rm(list = ls())
library(tidyverse)
source('../../R/strain_simulation_funs.R')
source('../../R/utility-functions.R')
```



## 1. Randomly place 5 antigens and 5 antibodies on a 2D map

* Coordinates are perfectly observed without measurement error

```{r}
set.seed(7)
n_antigens = 10
n_epitopes = 1
n_sera = 0
coords <- tibble(epitope = rep(c(1:n_epitopes), each = n_antigens),
                         antigen = rep(1:n_antigens, n_epitopes),
                         kind = 'antigen',
                         c1 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c2 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c3 = generate_ag_coords_random(n_antigens, n_epitopes)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))
n_dim = sum(grepl(pattern = 'c\\d', names(coords)))

## Shift so that epitope 1 of antigen 1 falls at the origin
shift_this_col = function(xx){ xx - xx[1] }
coords <- coords %>%
  mutate_at(.vars = vars(matches('c\\d?.+')), .funs = shift_this_col)
## Set all coords of ag2 epitope 1 to 0 other than c1, so that the ag falls on the x axis
modify_these_columns = grep(names(coords), pattern = 'c\\d?.+')[-1]
coords[2, modify_these_columns] = 0
if(coords[2,4]<0) coords[2,4]=-coords[2,4]
coords
```

### Plot coords

```{r}
cowplot::plot_grid(
ggplot(coords) +
  geom_point(aes(x = c1, y = c2)) +
  ggtitle('Hypothetical strains on a 3D map'),
ggplot(coords) +
  geom_point(aes(x = c1, y = c3))
,
ggplot(coords) +
  geom_point(aes(x = c2, y = c3)) 
)
```

## Calculate distances between coordinates

* In practices only ab-ag distances are observed, but for prelimiary testing we assume perfect observation (ab-ab and ag-ag distances are observed too).

```{r}
distances <- matrix(NA, n_antigens, n_antigens)
for(serum in 1:n_antigens){
  for(strain in 1:n_antigens){
    serum_coords = as.vector(coords[serum, 4:6])
    strain_coords = as.vector(coords[strain, 4:6])
    distances[serum, strain] = get_euclidean_distance(v1 = serum_coords, v2 = strain_coords)
  }
}
```



```{r}
fit_stan_MDS <- function(
  observed_distances, # vector of delay for each individual in the data
  n_antigens, 
  n_sera,
  n_dim,
  chains = 1, # Number of MCMC chains to run
  n_iter = 5000,
  cores = parallel::detectCores(logical = F),
  ...
) {
  library(rstan)
  
  model <- stan_model('../../stan/archive/MDS.stan')
  
  initfun <- function(){
  list(sigma = 1,
       ag2_c1 = runif(1, 0, 10),
       antigen_coords = matrix(runif((n_antigens-2)*n_dim, -10, 10), n_antigens-2, n_dim),
       serum_coords = matrix(runif(n_sera*n_dim, -10, 10), n_sera, n_dim)
  )
}
  
  model_input_data <- list(
    n_antigens = n_antigens,
    n_sera = n_sera,
    n_dim = n_dim,
    observed_distances = observed_distances
  )
  
  fit <- sampling(
    model, model_input_data, chains = chains, cores = cores, init=initfun,
    ...
  )
}

```

```{r}
fits <- fit_stan_MDS(observed_distances = distances, n_antigens = n_antigens, n_sera = n_antigens, n_dim = n_dim, chains = 1, n_iter = 5000)
```


## Compare the estimated and fitted distances to make sure they match
```{r}
compare_distances <- as.data.frame(summary(fits)$summary) %>%
  rownames_to_column(var = 'parameter') %>%
  filter(grepl(pattern = 'distance', parameter)) %>%
  tidyr::extract(parameter, into = c('antigen', 'serum'), regex = 'map_distances.(\\d?).(\\d?).', convert = T) %>%
  rename(estimated_distance = mean) %>%
  select(1:3) %>%
merge(
as.data.frame(distances) %>% 
  set_names(paste0(1:n_antigens)) %>%
  mutate(serum = 1:n_antigens) %>%
  pivot_longer(-serum, names_to = 'antigen', values_to = 'input_distance') %>%
  mutate(antigen = as.integer(antigen))
) 
```


```{r}
compare_distances %>%
  ggplot() +
  geom_point(aes(x = input_distance, y = estimated_distance))
```

```{r}
compare_distances %>%
  filter(abs(input_distance-estimated_distance)>1)
```
