---
title: "Test least squares methos"
output: html_notebook
---

Test the least squares method of Smith et al. 2004

```{r}
generate_alleles <- function(ndim, nalleles, range = 25){
  # Randomly draw coordinates in each dimension
  sapply(1:ndim, function(dim){runif(nalleles, 0, range)}) %>%
    as.tibble() %>%
    set_names(paste0('c', 1:ndim)) %>%
    # Assign allele ids
    mutate(allele = 1:nalleles)
}
```

```{r}
n_alleles = 5
## Generate coordinates for a separate set of antigens (Ag) and sera (Ab)
ab_df <- generate_alleles(2, n_alleles)
ag_df <- generate_alleles(2, n_alleles)

ab_coords <- as.matrix(ab_df %>% select(starts_with('c')))
ag_coords <- as.matrix(ag_df %>% select(starts_with('c')))

ag_ab_df<- bind_rows(list('ag' = ag_df,
               'ab' = ab_df),
          .id = 'kind') %>%
  rename(id = allele) %>%
  select(kind, id, starts_with('c'))

true_plot <- ag_ab_df %>%
  ggplot(aes(x = c1, y = c2, color = kind)) +
  geom_text(aes(label = id))

true_plot
```

```{r}
get_l2_distance<- function(v1, v2){
  stopifnot(length(v1) == length(v2))
  sqrt(sum((v1-v2)^2))
}

## Calculate the distance matrix
get_distances <- function(ab_coords, ag_coords){
  distmat = matrix(NA, nrow = nrow(ag_coords), ncol = nrow(ab_coords))
  rownames(distmat) = paste0('ag', 1:nrow(ag_coords))
  colnames(distmat) = paste0('ab', 1:nrow(ab_coords))
  for(ab in 1:nrow(ab_coords)){
    for(ag in 1:nrow(ag_coords)){
      distmat[ag, ab] = get_l2_distance(ab_coords[ab,], ag_coords[ag,])
    }
  }
  distmat
}
```

```{r}
observed_distances = get_distances(ab_coords, ag_coords)

## Test
sqrt(sum((ab_coords[3,]-ag_coords[5,])^2)) == observed_distances[5,3]
sqrt(sum((ab_coords[1,]-ag_coords[1,])^2)) == observed_distances[1,1]
sqrt(sum((ab_coords[2,]-ag_coords[5,])^2)) == observed_distances[5,2]
sqrt(sum((ab_coords[2,]-ag_coords[4,])^2)) == observed_distances[4,2]
sqrt(sum((ab_coords[2,]-ag_coords[3,])^2)) == observed_distances[3,2]
```


```{r}
fit_MDS_MLE <- function(
  observed_distances, # vector of delay for each individual in the data
  n_antigens,
  n_sera,
  n_dim
) {

  stopifnot(nrow(observed_distances) == n_antigens)
  stopifnot(ncol(observed_distances) == n_sera)
  
  loss_function <- function(pars, observed_distances, n_antigens, n_sera, n_dim){
    stopifnot(length(pars) == (n_antigens+n_sera)*n_dim)
    
    estimated_ag_coords <- matrix(pars[1:(n_antigens*n_dim)], n_antigens, n_dim)
    estimated_ab_coords <- matrix(pars[(n_antigens*n_dim+1):length(pars)], n_sera, n_dim)
    
    estimated_distances <- get_distances(estimated_ab_coords, estimated_ag_coords)
    sum((observed_distances - estimated_distances)^2)
  }

  generate_initial_guess <- function(n_antigens, n_sera, n_dim, lower_lim, upper_lim){
    ag_c1s <- runif(n_antigens, lower_lim, upper_lim); names(ag_c1s) = paste0('ag', 1:n_antigens, 'c1')
    ag_c2s <- runif(n_antigens, lower_lim, upper_lim); names(ag_c2s) = paste0('ag', 1:n_antigens, 'c2')
    ab_c1s <- runif(n_sera, lower_lim, upper_lim); names(ab_c1s) = paste0('ab', 1:n_antigens, 'c1')
    ab_c2s <- runif(n_sera, lower_lim, upper_lim); names(ab_c2s) = paste0('ab', 1:n_antigens, 'c2')
    
    c(ag_c1s, ag_c2s, ab_c1s, ab_c2s)
  }
  
  ## Repeat the fit 5 times from different initial values to verify convergence
  
  fit_list <- lapply(1:10, function(x) { optim(par = generate_initial_guess(n_antigens, n_sera, n_dim, lower_lim = 0, upper_lim = 50),
                                             fn = loss_function,
                                             observed_distances = observed_distances,
                                             n_antigens = n_antigens, 
                                             n_sera = n_sera,
                                             n_dim = n_dim,
                                             method = 'CG',
                                             control = list(maxit = 500))}
  )
  ## Check convergence
  if(!all(sapply(fit_list, function(ll) ll$convergence) == 0)){
    warning('Not all optimizations converged! - Check individual fits')
    return(fit_list)
  }
  ## Check that if all runs converged, that they all converged to the same solution
  errfun_value <- sapply(fit_list, function(ll) ll$value)
  if(!all(abs(errfun_value - mean(errfun_value) < .5))){warning('Not all runs converged to the same solution. Check individual fits.')}
  
  return(list(best_fit = fit_list[errfun_value == min(errfun_value)][[1]],
              all_fits = fit_list))
}

```


## Run the model to infer coordinates 
```{r}
fits <- fit_MDS_MLE(observed_distances = observed_distances, n_antigens = n_alleles, n_sera = n_alleles, n_dim = 2)
fits
```
## Visualize the difference between fits
```{r}
get_fits_df <- function(fit_obj){
  tibble(coordinate = fit_obj$par,
       label = names(fit_obj$par)) %>%
  tidyr::extract(label, into = c('kind','id', 'dim'), regex = '(a\\w)(\\d)c(\\d)', convert = T) %>%
  pivot_wider(names_from = dim, values_from = coordinate, names_prefix = 'c') 
}
```


Plot the solutions from each run.

Note that the runs with near-0 errfun all have very similar topology, whereas the higher-errfun solutions seem like a local min solution
```{r}
all_fits_df <- lapply(fits$all_fits, function(ll){ 
  get_fits_df(ll) %>%
  mutate(errfun = ll$value,
         convergence = ll$convergence) }) %>%
  bind_rows(.id = 'fit')
```

```{r}
all_fits_df %>%
  ggplot()+
  geom_point(aes(x = c1, y = c2, shape = kind, color = errfun)) +
  facet_wrap(.~fit)
```
Ok, so the coordinates are clearly not converging, but it looks like most of the maps with low error functions (dark points) are isometries of each other?



## Compare errorfunctions
```{r}
tibble(fit = 1:10,
       value = sapply(fits$all_fits, function(ll) ll$value)
) %>%
  ggplot() +
  geom_histogram(aes(x = value), bins = 100)
```



## Compare distance matrices
```{r}
fits_df <- get_fits_df(fits$best_fit)

get_distances(ab_coords = fits_df %>% dplyr::filter(kind == 'ab') %>% select(starts_with('c')) %>% as.matrix,
              ag_coords = fits_df %>% dplyr::filter(kind == 'ag') %>% select(starts_with('c')) %>% as.matrix)

observed_distances
```


## Compare distance matrices
```{r}
library(foreach)
all_fitted_distances = foreach(this_fit = 1:10) %do% {
  distmat = get_distances(ab_coords = all_fits_df %>% 
                  dplyr::filter(fit == this_fit) %>% 
                  dplyr::filter(kind == 'ab') %>% 
                  select(starts_with('c')) %>% as.matrix,
              ag_coords = all_fits_df %>% 
                  dplyr::filter(fit == this_fit) %>%
                  dplyr::filter(kind == 'ag') %>% 
                select(starts_with('c')) %>% 
                as.matrix)
  
  as_tibble(distmat) %>%
    mutate(antigen = 1:nrow(.)) %>%
    pivot_longer(-antigen, names_to = 'serum', names_prefix = 'ab', values_to = 'distance') %>%
    mutate(serum = as.integer(serum))
} %>%
  bind_rows(.id = 'fit') %>%
  merge(all_fits_df %>% select(fit, errfun), by = 'fit') %>%
  mutate(antigen_serum = paste0('antigen', antigen, '_serum', serum)) %>%
  mutate(`errfun<20` = errfun <20 ) 
```

## How much variance is there in the inferred distances from each run?
```{r}
all_fitted_distances %>%
  group_by(`errfun<20`, antigen, serum) %>%
  summarise(meandist = mean(distance),
            sem = sd(distance)/sqrt(n()))
```

very low sem values indicate that in all fits, the values converge to very similar distances



```{r}
## Boxplots of fitted distances for each ag-serum pair
all_fitted_distances %>%
  ggplot() +
  geom_boxplot(aes(x = antigen_serum, y = distance, color = `errfun<20`)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

Here, the tight boxplots (especially for runs with small error functions), we see that the estimated pairwise distances are consistent and reproducible, despite non-dentifiability of the antigen coordinates.

