---
title: "Multi epitope code tests"
output: html_notebook
---

Write a series of tests to make sure the multi epitope code is doing what we expect

```{r}
rm(list = ls())
library(tidyverse)
library(rstan)
extract <- tidyr::extract
filter <- dplyr::filter
summarise <- dplyr::summarise
source('../../R/utility-functions.R')
source('../../R/stan_funs_complete_data.R')
source('../../R/strain_simulation_funs.R')
source('../../R/rotate-shift-coords.R')

n_dim = 3
n_antigens = 15
n_ab = 0
n_chains = 4
n_iter = 5000 # half will be burn in
```


## 1. If there's only one epitope and if all Abs are exactly at the same position as the epitope, then we should recover the underlying map exactly

```{r}
set.seed(15)
antigen_coords <- tibble(epitope = rep(c(1:1), each = n_antigens),
                         antigen = rep(1:n_antigens, 1),
                         kind = 'antigen',
                         c1 = generate_ag_coords_random(n_antigens, 1),
                         c2 = generate_ag_coords_random(n_antigens, 1),
                         c3 = generate_ag_coords_random(n_antigens, 1)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))

shifted_coords <-  standardize_coordinates(coord_mat = antigen_coords %>% select(starts_with('c')) %>% as.matrix(), 
                                           ag1_row = 1, ag2_row = 2) %>%
  as_tibble() %>%
  set_names(c('c1', 'c2', 'c3'))

antigen_coords$c1 = shifted_coords$c1
antigen_coords$c2 = shifted_coords$c2
```


```{r}
plotly::plot_ly(antigen_coords, x = ~c1, y = ~c2, z = ~c3) %>% 
  plotly::add_markers() 
```

Generate ferret inputs
```{r}
generate_ferret_inputs(antigen_coords, 
                       relative_immunodominance = c(1), 
                       n_epitopes = 1, 
                       n_antigens = n_antigens, 
                       n_dim = n_dim, 
                       n_abs_per_serum = 10, 
                       sigma = .1, 
                       immunodominance_flag = 'E1', 
                       outdir = 'system_tests') 
E1_inputs <- read_rds('system_tests/3D_E1_immunodominance_inputs.rds')
```

Verify that calculated titers equal expected titers

(See equation 12 of the [Methods][https://www.overleaf.com/project/62152fe1d94828a32964742e].)
```{r}
E1_inputs$titer_map$epitope_level_titer = foreach(this_antigen = E1_inputs$titer_map$antigen,
        this_serum = E1_inputs$titer_map$serum,
        .combine = 'c') %do% {
          get_titer_epitope_level(ab_ag_coords = E1_inputs$ag_ab_coords, 
                                  this_antigen = this_antigen, 
                                  this_serum = this_serum, 
                                  this_epitope = 1, 
                                  alpha = .25, 
                                  r = 7)
        }

E1_inputs$titer_map %>% select(serum, antigen, titer, epitope_level_titer)
```

## Compare titer distances to Euclidean distances
```{r}
euclidean_distances <- get_ab_ag_distances(antigen_coords %>% select(starts_with('c')) %>% as.matrix(),
                                           antigen_coords %>% select(starts_with('c')) %>% as.matrix())

E1_inputs$titer_map %>%
  rowwise() %>%
  mutate(euclidean_distance = euclidean_distances[antigen, serum]) %>%
  ggplot() +
  geom_point(aes(x = titer_distance,
                 y = euclidean_distance)) +
  geom_abline(aes(intercept = 0, slope = 1))
```




## Titers are as expected

```{r}
ggplot(E1_inputs$titer_map) +
  geom_point(aes(x = titer, y = epitope_level_titer)) +
  geom_abline(aes(slope = 1, intercept = 0))
```




Infer titer map
```{r echo=FALSE, include=FALSE}
source('../../R/stan_funs_sparse_data.R')
```


## Split inputs into a test and training set
```{r}
split_inputs = test_train_split(E1_inputs$titer_map)
```

```{r}
distmat <- distance_matrix_format(split_inputs$train)
constrained_coords <- get_constrainted_coords(distmat = distmat, n_dim = n_dim, verbose = TRUE)
```

```{r}
E1_test = fit_stan_MDS(mod = '../../stan/MDS_predict_titer_sparse.stan',  
                       titer_map_train = split_inputs$train, 
                       titer_map_test = split_inputs$test, 
                       coord_prior_sd = 0.1, 
                       n_dim = n_dim, 
                       chains = 4, 
                       cores = 4, 
                       niter = 5000, 
                       diagnostic_file = NULL, 
                       debug = TRUE)
```

## Refit, selecting the best chain
```{r}
E1_test = fit_stan_MDS(mod = '../../stan/MDS_predict_titer_sparse.stan',  
                       titer_map_train = split_inputs$train, 
                       titer_map_test = split_inputs$test, 
                       coord_prior_sd = 0.1, 
                       n_dim = n_dim, 
                       chains = 4, 
                       cores = 4, 
                       niter = 5000, 
                       diagnostic_file = NULL, 
                       debug = FALSE)
```





In 3D, we're seeing a lack of convergence on the first run. Below it looks like a few of the chains are finding a lower-likelihood local max. If we rerun initializing with the best chain, we should converge.

```{r}
traceplot(E1_test, pars = c('lp__', names(E1_test)[21:25]))
```



```{r}
parnames = names(E1_test)
pairs(E1_test, pars = grep(pattern = 'coords', parnames, value = T)[1:5])
```

```{r}
parnames = names(E1_test)
pairs(E1_test, pars = grep(pattern = 'coords', parnames, value = T)[6:10])
```

```{r}
pairs(E1_test, pars = grep(pattern = 'coords', parnames, value = T)[11:16])
```

```{r}
pairs(E1_test, pars = grep(pattern = 'coords', parnames, value = T)[17:23])
```


```{r}
pairs(E1_test, pars = grep(pattern = 'coords', parnames, value = T)[24:30])
```


