---
title: "Multi epitope code tests"
output: html_notebook
---

Write a series of tests to make sure the multi epitope code is doing what we expect

```{r}
rm(list = ls())
library(tidyverse)
library(rstan)
extract <- tidyr::extract
filter <- dplyr::filter
summarise <- dplyr::summarise
source('../../R/utility-functions.R')
source('../../R/stan_funs_complete_data.R')
source('../../R/strain_simulation_funs.R')
source('../../R/rotate-shift-coords.R')

n_dim = 2
n_antigens = 5
n_ab = 0
n_chains = 4
n_iter = 5000 # half will be burn in
```


## 1. If there's only one epitope and if all Abs are exactly at the same position as the epitope, then we should recover the underlying map exactly

```{r}
set.seed(15)
antigen_coords <- tibble(epitope = rep(c(1:1), each = n_antigens),
                         antigen = rep(1:n_antigens, 1),
                         kind = 'antigen',
                         c1 = generate_ag_coords_random(n_antigens, 1),
                         c2 = generate_ag_coords_random(n_antigens, 1)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))

shifted_coords <-  standardize_coordinates(coord_mat = antigen_coords %>% select(starts_with('c')) %>% as.matrix(), ag1_row = 1, ag2_row = 2) %>%
  as_tibble() %>%
  set_names(c('c1', 'c2'))

antigen_coords$c1 = shifted_coords$c1
antigen_coords$c2 = shifted_coords$c2
```


```{r}
antigen_coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2))
```

Generate ferret inputs
```{r}
generate_ferret_inputs(antigen_coords, 
                       relative_immunodominance = c(1), 
                       n_epitopes = 1, 
                       n_antigens = n_antigens, 
                       n_dim = n_dim, 
                       n_abs_per_serum = 10, 
                       sigma = .1, 
                       immunodominance_flag = 'E1', 
                       outdir = 'system_tests') 
E1_inputs <- read_rds('system_tests/2D_E1_immunodominance_inputs.rds')
```

Verify that calculated titers equal expected titers

(See equation 12 of the [Methods][https://www.overleaf.com/project/62152fe1d94828a32964742e].)
```{r}
E1_inputs$titer_map$epitope_level_titer = foreach(this_antigen = E1_inputs$titer_map$antigen,
        this_serum = E1_inputs$titer_map$serum,
        .combine = 'c') %do% {
          get_titer_epitope_level(ab_ag_coords = E1_inputs$ag_ab_coords, 
                                  this_antigen = this_antigen, 
                                  this_serum = this_serum, 
                                  this_epitope = 1, 
                                  alpha = .25, 
                                  r = 7)
        }

E1_inputs$titer_map %>% select(serum, antigen, titer, epitope_level_titer)
```

## Titers are as expected

```{r}
ggplot(E1_inputs$titer_map) +
  geom_point(aes(x = titer, y = epitope_level_titer)) +
  geom_abline(aes(slope = 1, intercept = 0))
```

## Compare euclidean and titer distances
```{r}
euclidean_distances <- get_ab_ag_distances(antigen_coords %>% select(starts_with('c')) %>% as.matrix(),
                                           antigen_coords %>% select(starts_with('c')) %>% as.matrix())

E1_inputs$titer_map %>%
  rowwise() %>%
  mutate(euclidean_distance = euclidean_distances[antigen, serum]) %>%
  ggplot() +
  geom_point(aes(x = titer_distance,
                 y = euclidean_distance)) +
  geom_abline(aes(intercept = 0, slope = 1))
```

## Split inputs into a test and training set
```{r}
split_inputs <- test_train_split(E1_inputs$titer_map)
stopifnot(1:5 %in% split_inputs$train$antigen)
stopifnot(1:5 %in% split_inputs$train$serum)
```


Infer titer map
```{r echo=FALSE, include=FALSE}
source('../../R/stan_funs_sparse_data.R')
## Get the constraints based on the training set.
distmat <- distance_matrix_format(split_inputs$train)
constrained_coords <- get_constrainted_coords(distmat = distmat, n_dim = n_dim, verbose = F)
```

```{r}
E1_test = fit_stan_MDS(mod = '../../stan/MDS_predict_titer_sparse.stan', 
                       titer_map_train = split_inputs$train, 
                       titer_map_test = split_inputs$test,coord_prior_sd = 0.01, 
                       n_dim = n_dim, chains = 4, cores = 4, 
                       niter = 5000, diagnostic_file = NULL, debug = FALSE)
```


```{r}
E1_test
```



In 3D, we're seeing divergent transitions after warmup. Is this an identifiability issue?

```{r}
traceplot(E1_test, pars = 'lp__')
```



```{r}
parnames = names(E1_test)
pairs(E1_test, pars = grep(pattern = 'coords', parnames, value = T)[1:5])
```

```{r}
parnames = names(E1_test)
pairs(E1_test, pars = grep(pattern = 'coords', parnames, value = T)[6:10])
```

```{r}
pairs(E1_test, pars = grep(pattern = 'coords', parnames, value = T)[11:16])
```

```{r}
pairs(E1_test, pars = grep(pattern = 'coords', parnames, value = T)[17:20])
```





View the inferred points.

--> This looks good. The inferred points are an exact match, or a reflection about the x axis.

```{r}
extract_summary_coords(stan_fit = E1_test) %>%
  mutate(allele = as.factor(allele)) %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = allele, shape = kind))+
  geom_point(aes(x = c1, y = c2), data = antigen_coords, pch = 3)
```
Colors are inferred coordinates, and black crosses are the true coordinates.



Compare inferred and input titers
```{r}
get_titer_error(stan_fit = E1_test, 
                titer_map = split_inputs$train)[[2]] %>%
  ggplot() +
  geom_point(aes(x = mean_map_distance, y = titer_distance)) +
  geom_abline(aes(slope=1, intercept =0))
```


```{r}
get_predictive_errors(stan_fit = E1_test, 
                test_set = split_inputs$test)[[2]] %>%
  ggplot() +
  geom_point(aes(x = id, y = titer_error_med)) +
  geom_segment(aes(x = id, xend = id,  y = titer_error_0.025, yend = titer_error_0.975)) 
```
