---
title: "Muti epitope maps -- ferrets"
output: html_notebook
---

** This notebook modifies v0.0 so that we can use a single function to test differnt antigen maps and immunodominance levels**

Question --

When we use titers to map a multi-epitope pathogen, how does the inferred "strain" location relate to the locations/distances of individual epitopes?

**Workflow:**

1. Simulate a three-epitope pathogen with 5 antigens of each epitope represented on a 2D map.
2. Simulate ferret antibody repertoires to the epitope.
3. Calculate each ferret's titer to each strain to simulate a sero panel.
4. From the sero panel, infer the "strain map".
5. Repeat under conditions of even, skewed, and variable immunodominance

Qs:

* How do the inferred strain positions align with individual epitope positions?
* How does the relationship between strain and epitope position change as immunodominance becomes skewed?
* Can we infer the map consistently if ID varies between individuals?


```{r setup, include=FALSE}
rm(list = ls())
library(tidyr)
filter <- dplyr::filter
extract <- tidyr::extract
select <- dplyr::select
source('multi_epitope_code.R')
source('../code.R')
```

1. Simulate a 3-epitope pathogen with 5 antigens of each epitope.

```{r}
n_dim = 2
n_ag = 5
n_ab = 0
n_epitopes = 3
n_chains = 4
n_iter = 5000 # half will be burn in
relative_concentrations = c(1,1,1) # Relative immunodominance of each eptope (E-length vector). 1 is the null value. Value scales the concentration of all Abs to the corresponding epitope.

```

```{r}
generate_ag_coords <- function(n_ag, n_epitopes){
  lapply(1:n_epitopes, FUN = function(xx){
    cvec = numeric(n_ag)
    for(ii in 2:n_ag){
      cvec[ii] = cvec[ii-1] + rnorm(1, 2, 2)
    }
    return(cvec)
  }
  ) %>%
    unlist()
}
```


Draw antigen coords

```{r}
set.seed(21)
antigen_coords <- tibble(epitope = rep(c(1:n_epitopes), each = n_ag),
                         antigen = rep(1:n_ag, n_epitopes),
                         kind = 'antigen',
                         c1 = generate_ag_coords(n_ag, n_epitopes),
                         c2 = generate_ag_coords(n_ag, n_epitopes)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))
```
  
Transform so that E1 allele 2 falls on the x axis (this makes it slightl easier to interpret the inferred maps)

```{r}
shifted_coords <-  standardize_coordinates(coord_mat = antigen_coords %>% select(starts_with('c')) %>% as.matrix(), ag1_row = 1, ag2_row = 2) %>%
  as_tibble() %>%
  set_names(c('c1', 'c2'))

antigen_coords$c1 = shifted_coords$c1
antigen_coords$c2 = shifted_coords$c2
```


```{r}
antigen_coords %>%
  ggplot() +
  geom_line(aes(x=c1, y = c2, group = epitope), lty = 2, lwd = .1)+
  geom_point(aes(x = c1, y = c2, shape = antigen, color = epitope))+
  facet_wrap(.~epitope, labeller = label_both)
```

```{r}
even_immunodominance <- infer_ferret_map(antigen_coords, 
                relative_concentrations = c(1, 1, 1), 
                n_epitopes = 3, 
                n_antigens = 5, 
                n_abs_per_serum = 500, 
                plotdir = 'even_immunodominance', 
                sigma = .1)
```


```{r}
skewed_immunodominance <- infer_ferret_map(antigen_coords, 
                relative_concentrations = c(1+2/3, 1, 1/3), 
                n_epitopes = 3, 
                n_antigens = 5, 
                n_abs_per_serum = 500, 
                plotdir = 'skewed_immunodominance', 
                sigma = .1)
```


```{r}
E1_complete_immunodominance <- infer_ferret_map(antigen_coords, 
                relative_concentrations = c(3, 0, 0), 
                n_epitopes = 3, 
                n_antigens = 5, 
                n_abs_per_serum = 500, 
                plotdir = 'all_or_nothing_immunodominance', 
                sigma = .1)
```



## When immunodominance is even, differences in titer are smaller than differences between epitopes

* The inferred map is compressed (usually w.r.t. axis 2), relative to the centroid of the polygon defined by the corresponding epitopes.

```{r}
plot_inferred_original_map(antigen_coords,
                           even_immunodominance$stan_fits,
                           flip_these_chains_over_x = c(1)) + 
  ggtitle('even') + 
  theme(legend.position = 'none') + 
  facet_grid(allele~chain, labeller = label_both)
```

```{r}
plot_inferred_original_map(antigen_coords,
                           even_immunodominance$stan_fits,
                           flip_these_chains_over_x = c(1)) + 
  ggtitle('even') + 
  theme(legend.position = 'none') +
  facet_wrap(.~chain, labeller = label_both)
```
```{r}
plot_Ab_Ag_map(even_immunodominance$ag_ab_coords, antigen_coords) +
  ggtitle('Even immunodominance sera')
```


```{r}
even_plot = plot_inferred_original_map2(even_immunodominance$stan_fits, 
                            antigen_coords,
                            weights_for_shift = c(1,1,1),
                            flip_these_chains_over_x = c(1))
even_plot
```


```{r}
plot_compare_distances(even_immunodominance)
```

```{r}
plot_compare_distance_tiles(even_immunodominance)
```

```{r}
#centroid_counterfactual(even_immunodominance, c(1,1,1), antigen_coords)
```




## Even immunodominance summary:
The map's shape is qualitatively different from that of the centroids. Most inferred y coordinates are shrunk, relative to the y coordinate of the true centroid, but the ag4 inferred coords are an exception.


---------------------

## Skewed immunodominance plots

```{r}
plot_inferred_original_map(antigen_coords,
                           skewed_immunodominance$stan_fits,
                           flip_these_chains_over_x = c(1,3)) +
  ggtitle('skewed') + theme(legend.position = 'none')  + 
  facet_grid(allele~chain, labeller = label_both)
```

```{r}
plot_inferred_original_map(antigen_coords,
                           skewed_immunodominance$stan_fits,
                           flip_these_chains_over_x = c(1,3)) +
  ggtitle('skewed') + theme(legend.position = 'none')  + 
  facet_wrap(.~chain, labeller = label_both)
```

```{r}
plot_Ab_Ag_map(skewed_immunodominance$ag_ab_coords, antigen_coords) +
  ggtitle('Skewed immunodominance sera')
```

```{r}
skewed_plot = plot_inferred_original_map2(stan_fits = skewed_immunodominance$stan_fits,
                            antigen_coords = antigen_coords,
                            weights_for_shift = c(1+2/3, 1, 1/3),
                            flip_these_chains_over_x = c(1,3))
skewed_plot
```

```{r}
plot_compare_distances(skewed_immunodominance)
```


## Skewed summary:

Similarly, here we see that the inferred map has a different shape than the centroid map.
Most of the inferred coords are shrunk relative to the centroid map, but the shape of the map is also qualitatively different.

---------------------


## When a single epitope is immunodominant, we infer its coordinates exactly.

```{r}
plot_inferred_original_map(antigen_coords,
                           E1_complete_immunodominance$stan_fits,
                           flip_these_chains_over_x = 2:3) + 
  ggtitle('all or nothing') + theme(legend.position = 'none')  + 
  facet_grid(allele~chain, labeller = label_both) 
```

```{r}
plot_inferred_original_map(antigen_coords,
                           E1_complete_immunodominance$stan_fits,
                           flip_these_chains_over_x = 2:3) + 
  ggtitle('all or nothing') + theme(legend.position = 'none')  + 
  facet_wrap(.~chain, labeller = label_both) 
```
```{r}
plot_Ab_Ag_map(E1_complete_immunodominance$ag_ab_coords, antigen_coords)
```


```{r}
all_or_nothing_plot = plot_inferred_original_map2(stan_fits = E1_complete_immunodominance$stan_fits,
                            antigen_coords = antigen_coords,
                            weights_for_shift = c(1,0,0),
                            flip_these_chains_over_x = c(2,3)) 
all_or_nothing_plot
```

```{r}
plot_compare_distances(E1_complete_immunodominance)
```


## As expected, when E1 is totally dominant, we almost exactly recover the E1 coordinates.
```{r}
cowplot::plot_grid(even_plot + theme(legend.position = 'none') + ggtitle('even immunodominance'),
                  skewed_plot + theme(legend.position = 'none') + ggtitle('skewed_immunodominance'),
                  all_or_nothing_plot + theme(legend.position = 'none') + ggtitle('all or nothing E1 immunodominance'),
                  nrow = 1
)
ggsave('../plots/ferret_experiment_summary.png', width = 10, height = 4, units = 'in')
```


