---
title: "Muti epitope maps -- ferrets"
output: html_notebook
---

Question --

When we use titers to map a multi-epitope pathogen, how does the inferred "strain" location relate to the locations/distances of individual epitopes?

**Workflow:**

1. Simulate a three-epitope pathogen with 5 antigens of each epitope represented on a 2D map.
2. Simulate ferret antibody repertoires to the epitope.
3. Calculate each ferret's titer to each strain to simulate a sero panel.
4. From the sero panel, infer the "strain map".
5. Repeat under conditions of even, skewed, and variable immunodominance

Qs:

* How do the inferred strain positions align with individual epitope positions?
* How does the relationship between strain and epitope position change as immunodominance becomes skewed?
* Can we infer the map consistently if ID varies between individuals?


```{r setup, include=FALSE}
rm(list = ls())
library(tidyr)
filter <- dplyr::filter
extract <- tidyr::extract
select <- dplyr::select
source('../code.R')
```

1. Simulate a 3-epitope pathogen with 5 antigens of each epitope.

```{r}
n_dim = 2
n_ag = 5
n_ab = 0
n_epitopes = 3
n_chains = 4
relative_concentrations = c(1,1,1) # Relative immunodominance of each eptope (E-length vector). 1 is the null value. Value scales the concentration of all Abs to the corresponding epitope.

```

```{r}
generate_ag_coords <- function(n_ag, n_epitopes){
  lapply(1:n_epitopes, FUN = function(xx){
    cvec = numeric(n_ag)
    for(ii in 2:n_ag){
      cvec[ii] = cvec[ii-1] + rnorm(1, 2, 2)
    }
    return(cvec)
  }
  ) %>%
    unlist()
}

antigen_coords <- tibble(epitope = rep(c(1:n_epitopes), each = n_ag),
                         antigen = rep(1:n_ag, n_epitopes),
                         kind = 'antigen',
                         c1 = generate_ag_coords(n_ag, n_epitopes),
                         c2 = generate_ag_coords(n_ag, n_epitopes)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))
```
  


```{r}
antigen_coords %>%
  ggplot() +
  geom_line(aes(x=c1, y = c2), lty = 2, lwd = .1)+
  geom_point(aes(x = c1, y = c2, shape = antigen, color = epitope))+
  facet_wrap(.~epitope, labeller = label_both)
```


2. Simulate ferret antibody repertoires to the epitope.
```{r}
generate_ferret_repertoire <- function(native_epitope_coords, ## A data frame with columns for epitope, strain id, and each coordinate
                                       n_epitopes,
                                       n_ab = 99,
                                       rel_immuno){
  ## Figure out how many Abs to draw for each epitope
  draw_this_many = floor(n_ab*rel_immuno/sum(rel_immuno))
  remainder = n_ab - sum(draw_this_many)
  cat(sprintf('Drawing %i Abs per individual. %i were requested. Remainder is %i.\n', sum(draw_this_many), n_ab, remainder))
    
  long_coords = select(native_epitope_coords, 'epitope', starts_with('c')) %>% pivot_longer(starts_with('c'), values_to = 'native_coord')

  mapply(FUN = function(nn, xx){
    rnorm(nn, mean = xx, sd = 1)
  },
  nn = draw_this_many,
  xx = pull(long_coords, native_coord)) %>%
    t() %>%
    as.tibble() %>%
  bind_cols(long_coords) %>%
    pivot_longer(starts_with('V'), names_to = 'antibody', names_prefix = 'V') %>%
    ungroup() %>%
    pivot_wider(names_from = 'name', values_from = 'value', id_cols = c('epitope', 'antibody'))
  }
```

## Apply for each strain
```{r}
ferret_repertoires <- lapply(1:n_ag, function(this_ag){
  generate_ferret_repertoire(native_epitope_coords = antigen_coords %>% filter(antigen == this_ag), 
                             n_epitopes = n_epitopes,
                             n_ab = 100, 
                             rel_immuno = c(1,1,1))
}) %>%
  bind_rows(.id = 'antigen')
```

```{r}
## visualize

ferret_repertoires %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = epitope)) +
  geom_point(data = antigen_coords, aes(x = c1, y = c2))+
  facet_wrap(.~antigen)

## Colored dots are Abs, and black dots are epitope coords
```



3. Calculate each ferret's titer to each strain to simulate a sero panel.
```{r}
## Merge the Ab and Ag data frames
merged_df <- merge(antigen_coords %>%
        select(epitope, antigen, c1, c2),
      ferret_repertoires,
      by = c('epitope', 'antigen'), 
      suffixes = c('_Ag', '_Ab'))

```

## Write a function to calculate the titer from the Ab repertoire
```{r}
extract_titer_inputs <- function(antigen_id, # To which antigen are we calculating titer 
                                 serum_id, # Which serum are we using?
                                 merged_df # data frame of antigen and antibody locations
                                 ){
  
  ## Extract ab_position_list
  these_serum_abs <- merged_df %>% # Subset the master data frame
    filter(antigen == serum_id) %>%
    select(epitope, matches('c\\d_Ab'))
  
  ab_position_list = lapply(unique(these_serum_abs$epitope), ## Reformat into a list of matrices, one for each epitope
                            function(this.epitope){
                              these_serum_abs %>% 
                                filter(epitope == this.epitope) %>%
                                select(starts_with('c')) %>%
                                as.matrix
                            }) %>%
    set_names(paste0('E', 1:length(.)))
  
  
  ## Extract the ag coordinates as a matrix
  this_ag = merged_df %>%
    filter(antigen == antigen_id) %>%
    select(epitope, matches('c\\d_Ag')) %>%
    distinct() %>%
    select(-epitope) %>%
    as.matrix()
  ag_list = lapply(1:nrow(this_ag), function(ii){
    this_ag[ii,]
  })
  
  ## Return
  list(ab_position_list = ab_position_list, 
       ag_list = ag_list)
}


solve_for_titer_multi_epitope <- function(ab_position_list, # A list of E matrices whose rows represent the coordinates of Abs specific to epitope E. Each matrix should be of dimension  [n_E, d].
                                          relative_concentrations = NULL, # An optional vector that scales the immunodominance of each epitope. (Assume in eq. 18 that concentration is proportional to the cognate epitope's immunodominance)
                                          ag_list, # A list of length E giving the coordinate vectors of each epitope within the antigen
                                          alpha, # Ab potency (equal and fixed across all Abs)
                                          r # Maximum log2 titer
                                          ){ # Location of each epitope in the antigen. Vector of length E.

   # cat(sprintf('ag list length is %i; ab list lenght is %i', length(ag_list), length(ab_position_list)))
    stopifnot(length(ag_list)==length(ab_position_list))
    if(length(relative_concentrations) < 1) relative_concentrations = rep(1, length(ab_position_list))
    stopifnot(length(relative_concentrations) == length(ab_position_list))


    titer_fun <- function(z, ab_position_list, x_ag, alpha, r, relative_concentrations){ 

      # Each eptiope is a factor in Eq 13. This function calculates those factors
      one_factor <- function(ab_positions, # matrix of Ab positions to one epitope
                             x_ag = x_ag, # vector of epitope's position
                             z, # titer (we'll solve for this later)
                             alpha, # potency
                             r,  # max titer
                             this_immunodominance # the immunodominance (relative concentration) of this epitope
                             ){
       # cat(sprintf(' ID is %2.1f', this_immunodominance))
        ab_distances = apply(ab_positions, 1, FUN = function(v1){
          get_euclidean_distance(v1, v2 = x_ag)
        })
        
        ab_affinities = 2^(r-ab_distances) # Eq 6
        scaled_affinities = ab_affinities*this_immunodominance
        this_factor = (1+alpha/z*sum(scaled_affinities))/(1+1/z*sum(scaled_affinities)) # Eq 13
        return(this_factor)
        
      }
        ## This function calculates each factor and takes the product to solve for titer
        factors<-mapply(FUN = one_factor, ab_positions = ab_position_list, x_ag = ag_list, this_immunodominance = relative_concentrations, z = z, alpha = alpha, r = r)  
        #cat(print('\n'))
        # cat(print('factors are\n'))
        # print(factors)
        prod(factors) - 1/2
    }

    ## Solve for titer and return
    numerical_soln <- uniroot(f = titer_fun, 
            interval = c(.0000000001, 10^10), 
            ab_position_list = ab_position_list, 
            x_ag = ag_list, 
            alpha = alpha, 
            r = r,
            relative_concentrations = relative_concentrations,
            tol =  .Machine$double.eps^.5)       

    ## Verify that the solution returns 0
    stopifnot(zeroish(titer_fun(numerical_soln$root, ab_position_list, x_ag, alpha, r, relative_concentrations)))

    return(numerical_soln$root)
}
```

## Test the titer function

test.1. Titer to ag 1 should decrease from serum 1 to 5:
```{r}
lapply(1:n_ag, function(this.serum){
  with(extract_titer_inputs(antigen_id = 1, serum_id = this.serum, merged_df = merged_df),
       solve_for_titer_multi_epitope(ab_position_list = ab_position_list, 
                                     relative_concentrations = relative_concentrations, 
                                     ag_list = ag_list, 
                                     alpha = .25,
                                     r = 7)
  )
}
)

```
Pass.



test.2. If we make Abs to epitope 1 perfectly matched, and Abs to all other epitopes mismatched:

2.a Titer should increase as epitope 1 becomes more immunodominant
2.b Titer should be equal to the single-epitope titer (E1. 16) when epitope 1 is fully immunodominant
```{r}

## This function replaces all Abs to epitope 1 with the exact epitope 1 coords
get_input_list <- function(this.serum){
  input_list = extract_titer_inputs(antigen_id = 2, serum_id = this.serum, merged_df = merged_df)
  input_list$ab_position_list[[1]] = matrix(input_list$ag_list[[1]], nrow = nrow(input_list$ab_position_list[[1]]), ncol = ncol(input_list$ab_position_list[[1]]), byrow = T)
  # Add mismatch to other Abs
  input_list$ab_position_list[[2]] = input_list$ab_position_list[[2]]+1
  input_list$ab_position_list[[3]] = input_list$ab_position_list[[3]]+1
  input_list
}

input_list = get_input_list(this.serum = 2)
ab_position_list = input_list$ab_position_list
ag_list = input_list$ag_list


cat(print('Even immunodominance titer:\n'))
## Titer when ID is even
with(get_input_list(this.serum = 2),
       solve_for_titer_multi_epitope(ab_position_list = ab_position_list, 
                                     relative_concentrations = c(1,1,1), 
                                     ag_list = ag_list, 
                                     alpha = .25,
                                     r = 7))

cat(print('Skewed immunodominance titer should be > even titer:\n'))
## Titer when ID is skewed (should be higher b/c epitope 1 Abs are perfectly matched)
with(get_input_list(this.serum = 2),
       solve_for_titer_multi_epitope(ab_position_list = ab_position_list, 
                                     relative_concentrations = c(2, .5, .5), 
                                     ag_list = ag_list, 
                                     alpha = .25,
                                     r = 7))

cat(print('Complete E1 immunodominance titer should be > skewed titer:\n'))
## Titer when ID of epitope 1 is perfect
with(get_input_list(this.serum = 2),
       solve_for_titer_multi_epitope(ab_position_list = ab_position_list, 
                                     relative_concentrations = c(3, 0, 0), 
                                     ag_list = ag_list, 
                                     alpha = .25,
                                     r = 7))

cat(print('Single epitope titer should be = Complete E1 dominant titer:\n'))
## Check against the single-epitope titer
distances = apply(ab_position_list[[1]], MARGIN = 1, FUN = function(this.ab){get_euclidean_distance(v1 = this.ab, v2 = ag_list[[1]])})
affinities = 2^(7-distances)
(1-2*.25)*sum(3*affinities)
```

Pass.


## Calculate the titer panel values
```{r}
## Wrapper to calculate the titer for a given serum and antigen
titer_wrapper <- function(this.serum, 
                          this.antigen){
  titer = with(extract_titer_inputs(antigen_id = this.antigen, 
                            serum_id = this.serum, 
                            merged_df = merged_df),
       solve_for_titer_multi_epitope(ab_position_list = ab_position_list, 
                                     relative_concentrations = relative_concentrations, 
                                     ag_list = ag_list, 
                                     alpha = .25,
                                     r = 7)
  )
  c('serum' = this.serum, 'antigen' = this.antigen, 'titer' = titer)
}
```


```{r}
titer_map <- with(expand.grid(antigen = 1:n_ag, serum = 1:n_ag),
     mapply(FUN = titer_wrapper, 
            this.serum = serum,
            this.antigen = antigen)) %>% 
  t() %>%
  as_tibble() %>%
  set_names(c('serum', 'antigen', 'titer')) %>%
  mutate(logtiter = log2(titer/10)) %>%
  group_by(serum) %>%
  mutate(serum_potency = max(logtiter)) %>%
  ungroup() %>% group_by(antigen) %>%
  mutate(antigen_avidity = max(logtiter)) %>%
  ungroup() %>%
  mutate(titer_distance = (serum_potency+antigen_avidity)/2 - logtiter)
```


```{r}
ggplot(titer_map) +
  geom_point(aes(x = serum, y = antigen, color = titer_distance)) +
  scale_color_viridis_c()
```


4. From the sero panel, infer the "strain map".

```{r}

format_stan_inputs <- function(titer_map){
  antigens = unique(titer_map$antigen)
  sera = unique(titer_map$serum)
  
  distmat = matrix(NA, length(antigens), length(sera))
  for(aa in 1:length(antigens)){
    for(ss in 1:length(sera)){
      distmat[aa,ss] = filter(titer_map, antigen == aa & serum == ss) %>%
        pull(titer_distance)
    }
  }
  distmat
}

```

## test
```{r}
format_stan_inputs(titer_map)
```


```{r}
fit_stan_MDS <- function(
  mod = 'MDS.stan',
  observed_distances, # n_antigen x n_antibody matrix of distances
  n_antigens, # integer
  n_sera, # integer
  n_dim, # integer
  chains = 3, # Number of MCMC chains to run
  cores = parallel::detectCores(logical = F), # For the cluster
  ...
) {
  library(rstan)
  
  stopifnot(nrow(observed_distances) == n_antigens)
  stopifnot(ncol(observed_distances) == n_sera)
  stopifnot(n_antigens>=3)
  
  model <- stan_model('../Bayesian_stan/MDS.stan')
  (print(model))
  
  model_input_data <- list(
    n_strains = n_antigens,
    n_sera = n_sera,
    n_dim = n_dim,
    observed_distances = observed_distances
  )
  
    initfun <- function(){
    list(sigma = 1,
         ag2_c1 = runif(1, 0, 10),
         strain_coords = matrix(runif((n_antigens-2)*n_dim, -10, 10), n_antigens-2, n_dim),
         serum_coords =  matrix(runif(n_sera*n_dim, -10, 10), n_sera, n_dim)
         )
    }
  
  fit <- sampling(
    model, model_input_data, chains = chains, cores = cores, 
    init = initfun,
    iter = 5000,
    control = list(adapt_delta = 0.89,
                    max_treedepth = 14),
    ...
  )
}

```

```{r}
fits <- fit_stan_MDS(observed_distances = format_stan_inputs(titer_map), 
                     n_antigens = n_ag, 
                     n_sera = n_ag, 
                     n_dim = n_dim, 
                     chains = n_chains)
```

## Visualize trace plots
```{r}
rstan::traceplot(fits, pars = names(fits))
```

## Visualize the pairs plots
```{r}
pairs(fits, pars = names(fits)[1:5])
```

## Check the distance matrix
```{r}
raw_fits <- rstan::extract(fits)

long_antigen_coords <- lapply(1:(n_ag-2), function(ll) raw_fits$antigen_coords[,ll,] %>%
                                                      magrittr::set_colnames(c('c1', 'c2')) %>%
                           as.tibble() %>%
                           mutate(chain = rep(1:n_chains, each = 2500),
                                  id = ll+2,
                                  iter = 1:nrow(.),
                                  kind = 'antigen')) %>%
                    bind_rows()

long_serum_coords <- lapply(1:n_ag, function(ll) raw_fits$serum_coords[,ll,] %>%
                                                      magrittr::set_colnames(c('c1', 'c2')) %>%
                           as.tibble() %>%
                           mutate(chain = rep(1:n_chains, each = 2500),
                                  id = ll,
                                  iter = 1:nrow(.),
                                  kind = 'serum')) %>%
                    bind_rows()

ag2_summary <- tibble(id = 2, kind = 'ag2', chain = rep(1:n_chains, each = 2500), c1 = raw_fits$ag2_c1) %>%
  group_by(id, kind, chain) %>%
  summarise(c1.10 = quantile(c1, .1),
            c1 = median(c1),
            c1.90 = quantile(c1, .9)) %>%
  mutate(c2.10 = NA, c2 = 0, c2.90 = NA)

summary_coords <- bind_rows(long_antigen_coords,
          long_serum_coords) %>%
  group_by(id, kind, chain) %>%
  summarise(c1.10 = quantile(c1, .1),
            c1 = median(c1),
            c1.90 = quantile(c1, .9),
            c2.10 = quantile(c2, .1),
            c2 = median(c2),
            c2.90 = quantile(c2, .9)) %>%
  ungroup() %>%
  bind_rows(tibble(id = 1,
                   kind = 'fixed_ag1', chain = 1:3, c1.10 = NA, c1 = 0, c1.90 = NA, c2.10 = NA, c2 = 0, c2.90 = NA)) %>%
    bind_rows(ag2_summary)
```

## Plot
```{r}
reformatted_df <- antigen_coords %>%
               mutate(kind = paste0('E', epitope)) %>%
               rename(id = antigen) %>%
               select(id, kind, starts_with('c'))
                      
summary_coords %>%
  mutate(id = as.factor(id)) %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, shape = kind, color = id), data = reformatted_df) +
  geom_point(aes(x = c1, y = c2, shape = kind, fill = id)) +
facet_wrap(.~chain,  ncol = 2, nrow = 2) +
  scale_shape_manual(values = c(15, 21, 2, 3, 4, 17, 23)) +
  ggtitle('Even immunodominance')

```

