---
title: "Bayesian MDS in Stan"
output: html_notebook
---

Tryfixing one point to resolve the tranformation issue.

```{r}
generate_alleles <- function(ndim, nalleles, range = 10){
  # Randomly draw coordinates in each dimension
  sapply(1:ndim, function(dim){runif(nalleles, -range/2, range)}) %>%
    as.tibble() %>%
    set_names(paste0('c', 1:ndim)) %>%
    # Assign allele ids
    mutate(id = 1:nalleles)
}
```

```{r}
## Generate coordinates for a separate set of antigens (Ag) and sera (Ab)
ab_df <- generate_alleles(2, 5)
ag_df <- generate_alleles(2, 5)

ab_coords <- as.matrix(ab_df %>% select(starts_with('c')))
ag_coords <- as.matrix(ag_df %>% select(starts_with('c')))

ag_ab_df<- bind_rows(list('ag' = ag_df,
               'ab' = ab_df),
          .id = 'kind') %>%
  select(kind, id, starts_with('c'))

true_plot <- ag_ab_df %>%
  ggplot(aes(x = c1, y = c2, color = kind)) +
  geom_text(aes(label = id))

true_plot
```

```{r}
get_l2_distance<- function(v1, v2){
  stopifnot(length(v1) == length(v2))
  sqrt(sum((v1-v2)^2))
}

## Calculate the distance matrix
get_distances <- function(ab_coords, ag_coords){
  distmat = matrix(NA, nrow = nrow(ag_coords), ncol = nrow(ab_coords))
  rownames(distmat) = paste0('ag', 1:nrow(ag_coords))
  colnames(distmat) = paste0('ab', 1:nrow(ab_coords))
  for(ab in 1:nrow(ab_coords)){
    for(ag in 1:nrow(ag_coords)){
      distmat[ag, ab] = get_l2_distance(ab_coords[ab,], ag_coords[ag,])
    }
  }
  distmat
}
```

```{r}
observed_distances = get_distances(ab_coords, ag_coords)

## Test
# sqrt(sum((ab_coords[3,]-ag_coords[5,])^2)) == observed_distances[5,3]
# sqrt(sum((ab_coords[1,]-ag_coords[1,])^2)) == observed_distances[1,1]
# sqrt(sum((ab_coords[2,]-ag_coords[5,])^2)) == observed_distances[5,2]
# sqrt(sum((ab_coords[2,]-ag_coords[4,])^2)) == observed_distances[4,2]
# sqrt(sum((ab_coords[2,]-ag_coords[3,])^2)) == observed_distances[3,2]
```


```{r}
fit_MDS_MLE_fix1 <- function(
  observed_distances, # vector of delay for each individual in the data
  n_antigens,
  n_sera,
  n_dim,
  ag_1_coords
) {

  stopifnot(nrow(observed_distances) == n_antigens)
  stopifnot(ncol(observed_distances) == n_sera)
  
  loss_function <- function(pars, observed_distances, n_antigens, n_sera, n_dim){
    stopifnot(length(pars) == ((n_antigens-1)+n_sera)*n_dim)
    
    estimated_ag_coords <- rbind(ag_1_coords, matrix(pars[1:((n_antigens-1)*n_dim)], n_antigens-1, n_dim)) %>% as.matrix()
    estimated_ab_coords <- matrix(pars[((n_antigens-1)*n_dim+1):length(pars)], n_sera, n_dim)
    
    estimated_distances <- get_distances(estimated_ab_coords, estimated_ag_coords)
    sum((observed_distances - estimated_distances)^2)
  }

  generate_initial_guess <- function(n_antigens, n_sera, n_dim){
    ag_c1s <- runif(n_antigens-1, -10, 10); names(ag_c1s) = paste0('ag', 2:n_antigens, 'c1')
    ag_c2s <- runif(n_antigens-1, -10, 10); names(ag_c2s) = paste0('ag', 2:n_antigens, 'c2')
    ab_c1s <- runif(n_sera, -10, 10); names(ab_c1s) = paste0('ab', 1:n_sera, 'c1')
    ab_c2s <- runif(n_sera, -10, 10); names(ab_c2s) = paste0('ab', 1:n_sera, 'c2')
    
    c(ag_c1s, ag_c2s, ab_c1s, ab_c2s)
  }
  
  ## Repeat the fit 5 times from different initial values to verify convergence
  
  fit_list <- lapply(1:10, function(x) { optim(par = generate_initial_guess(n_antigens, n_sera, n_dim),
                                             fn = loss_function,
                                             observed_distances = observed_distances,
                                             n_antigens = n_antigens, 
                                             n_sera = n_sera,
                                             n_dim = n_dim,
                                             method = 'CG',
                                             control = list(maxit = 500))}
  )
  ## Check convergence
  if(!all(sapply(fit_list, function(ll) ll$convergence) == 0)){
    warning('Not all optimizations converged! - Check individual fits')
  }
  ## Check that if all runs converged, that they all converged to the same solution
  errfun_value <- sapply(fit_list, function(ll) ll$value)
  if(!all(abs(errfun_value - mean(errfun_value) < .5))){warning('Not all runs converged to the same solution. Check individual fits.')}
  
  return(list(best_fit = fit_list[errfun_value == min(errfun_value)][[1]],
              all_fits = fit_list))
}

```

```{r}
fits <- fit_MDS_MLE_fix1(observed_distances = observed_distances, n_antigens = 5, n_sera = 5, n_dim = 2, ag_1_coords= ag_coords[1,])
fits
```
## Visualize the difference between fits
```{r}
get_fits_df <- function(fit_obj){
  tibble(coordinate = fit_obj$par,
       label = names(fit_obj$par)) %>%
  tidyr::extract(label, into = c('kind','id', 'dim'), regex = '(a\\w)(\\d)c(\\d)', convert = T) %>%
  pivot_wider(names_from = dim, values_from = coordinate, names_prefix = 'c') %>%
  bind_rows(ag_df[1,] %>% mutate(kind = 'ag'))
}
```


Plot the solutions from each run.

Note that the runs with near-0 errfun all have very similar topology, whereas the higher-errfun solutions seem like a local min solution
```{r}
all_fits_df <- lapply(fits$all_fits, function(ll){ 
  get_fits_df(ll) %>%
  mutate(errfun = ll$value,
         convergence = ll$convergence) }) %>%
  bind_rows(.id = 'fit')
```

```{r}
all_fits_df %>%
  mutate(kind = ifelse(kind == 'ag' & id == 1, 'fixed', kind)) %>%
  ggplot()+
  geom_point(aes(x = c1, y = c2, shape = kind, color = errfun)) +
  facet_wrap(.~fit)
```


## Compare errorfunctions
```{r}
tibble(fit = 1:10,
       value = sapply(fits$all_fits, function(ll) ll$value)
) %>%
  ggplot() +
  geom_histogram(aes(x = value), bins = 100)
```

## Plot
The mds is an isometry of the true coordinates
```{r}
fits_df <- get_fits_df(fits$best_fit) %>%
  arrange(kind, id)

mds_plot <- fits_df %>%
  ggplot(aes(x = c1, y = c2, color = kind))+
  geom_text(aes(label = id))

cowplot::plot_grid(
  true_plot+ggtitle('truth'),
  mds_plot+ggtitle('mds'))
```


## We should be able to reflect and rotate the solutions to obtain a good comparison to the original
```{r}
align_mds_with_original <- function(mds_df, original_df){
  mds_coords = all_fits_df %>% dplyr::filter(fit == '2') %>% arrange(desc(kind), id) %>% select(starts_with('c')) %>% select(-convergence) %>% as.matrix()
  original_coords = original_df %>% arrange(desc(kind), id) %>% select(starts_with('c')) %>% as.matrix()
  
  ## Shift so that the ag1 coordinate is at the origin for both maps
  shifted_mds = apply(mds_coords, 2, function(vv) vv-vv[1])
  shifted_original = apply(original_coords, 2, function(vv) vv-vv[1])

  ## Get the angles between corresponding coordinates in each map
  get_cosine <- function(v1, v2){
    dot_prod = sum(v1*v2) # dot product
    mag_v1 = sqrt(sum(v1^2))
    mag_v2 = sqrt(sum(v2^2))
    cos_theta = dot_prod/(mag_v1*mag_v2)
    theta = acos(cos_theta)
    theta
  }
  
  angle_original_to_mds = sapply(2:nrow(shifted_original), function(ii) get_cosine(shifted_original[ii,], shifted_mds[ii,]))
  ## If all the angles are equal, then the mds map is a rotation of the original
  ## If the angles are different, then the mds map is a reflection and a rotation
  if(!all( abs(angle_original_to_mds - mean(angle_original_to_mds)) < .1 )){
    ## Reflect across a vertical axis
    shifted_mds[,1] = shifted_mds[,1] - min(shifted_mds[,1]) # Shift so that all points are non-negative on the x-axis
    reflected_mds = t(matrix(c(-1, 0, 0, 1), 2, 2) %*% t(shifted_mds)) # Reflect across the y-axis
    shifted_mds = apply(reflected_mds, 2, function(vv) vv-vv[1]) # Re-shift the ag 1 point to the origin
    angle_original_to_mds = sapply(2:nrow(shifted_original), function(ii) get_cosine(shifted_original[ii,], shifted_mds[ii,])) ## Recalculate angles
    stopifnot(all( abs(angle_original_to_mds - mean(angle_original_to_mds)) < .1 )) ## Break if the reflection was not successful
  }
  
  ## Rotate to align with the original
  theta = 2*pi - mean(angle_original_to_mds) # Get the angle
  V = matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), nrow = 2, ncol = 2) # Get the rotation matrix
  rotated_mds = t(V %*% t(shifted_mds)) # Rotate by theta
  
  ## Shift to align with the original and return
  aligned_mds = sapply(1:ncol(rotated_mds), function(cc) rotated_mds[,cc] - rotated_mds[1,cc] +  original_coords[1,cc]) 
  original_df %>%
    mutate(mds1 = mds_coords[,1],
           mds2 = mds_coords[,2],
           aligned_mds1 = aligned_mds[,1],
           aligned_mds2 = aligned_mds[,2])
}
```


## ALign mds with original coords
```{r}
check_mds <- align_mds_with_original(mds_df = fits_df,
                        original_df = ag_ab_df) 

check_mds %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2), pch = 16) +
  geom_point(aes(x = aligned_mds1, y = aligned_mds2, color = kind), pch = 9)

```

## It works!