---
title: "Muti epitope maps -- ferrets"
output: html_notebook
---


## Goal - build intuition for what we expect when we infer the map assuming even immunodominance

**Approach**

1. Define distance measures in which we can predict easily what the map will look like
2. Study how the titer distance chanes in comparison to these intuitive measures

**Metrics**

1. Weighted average - $d^{map}_{ij} = \sum_e d_{ij}^{(e)}w_e$, where $\sum_e w_e = 1$, and $w_e$ represents immunodominance.
    
    * Expectation - map will reflect the weighted epitope-specific centroids.
    * Map error will be minimal.
    
2. Min distances - $d^{map}_{ij} = min(d_{ij}^{(e)})$

    * Expectation - map will reflect the max pairwise distances between epitopes.
    * Map error will be higher -- it's not clear that there's an exact solution here
    
3. Actual titer model -- 
  
    How does titer distance relate to average distnace, or to max distance?



```{r}
source('../../multi_epitope_code.R')
source('../../code.R')
library('foreach')


n_antigens = 20
n_ab = 0
n_chains = 4
n_iter = 5000 # half will be burn in
n_epitopes = 3

outdir = 'ferret_immunodominance_tests'
if(!dir.exists(outdir)) dir.create(outdir)
```

## Generate antigen coordinates. Each antigen consists of 3 epitopes.
```{r}
set.seed(13)
antigen_coords <- tibble(epitope = rep(c(1:n_epitopes), each = n_antigens),
                         antigen = rep(1:n_antigens, n_epitopes),
                         kind = 'antigen',
                         c1 = generate_ag_coords_random(n_antigens, n_epitopes),
                         c2 = generate_ag_coords_random(n_antigens, n_epitopes)) %>%
  mutate(epitope = as.factor(epitope),
         antigen = as.factor(antigen))


## Standardize so that Ag1 epitope 1 is at the origin and Ag2 epitope 1 is on the x-axis
shifted_coords <-  standardize_coordinates(coord_mat = antigen_coords %>% select(starts_with('c')) %>% as.matrix(), ag1_row = 1, ag2_row = 2) %>%
  as_tibble() %>%
  set_names(c('c1', 'c2'))

antigen_coords$c1 = shifted_coords$c1
antigen_coords$c2 = shifted_coords$c2

write_csv(antigen_coords, paste0(outdir, '/antigen_coords.csv'))
```

## Plot the coordinates
```{r}
antigen_coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), alpha = .2)
ggsave(filename = paste0(outdir, '/epitope_coords_polygon.png'))
```

```{r}
antigen_coords %>%
  ggplot() +
  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  facet_wrap(.~epitope)
ggsave(filename = paste0(outdir, '/epitope_coords_grid.png'))
```

## Weighted average distance matrices
```{r}
get_one_epitope_distances <- function(this.epitope){
  this.coordmat = antigen_coords  %>% filter(epitope == this.epitope) %>% select(starts_with('c')) %>% as.matrix()
  get_ab_ag_distances(ab_coords = this.coordmat,
                                            ag_coords = this.coordmat) %>%
    as.tibble() %>%
    mutate(epitope = this.epitope, 
           antigen = 1:nrow(.)) %>%
    pivot_longer(-c(antigen, epitope), names_to = 'serum', names_prefix = 'ab', values_to = 'distance') %>%
    mutate(serum = as.integer(serum))
}

epitope_distances <- lapply(1:n_epitopes, get_one_epitope_distances) %>% bind_rows()
```


```{r}
weighted_avg_distance_map <- epitope_distances %>%
  group_by(antigen, serum) %>%
  summarise(mean_distance = mean(distance),
            skewed_weighted_distance = sum(distance * c(1+2/3, 1, 1/3)/3),
            max_distance = max(distance),
            min_distance = min(distance),
            dist_variance = var(distance))
weighted_avg_distance_map
```


## Generate a serum with one Ab per epitope
```{r}
simple_serum = generate_ferret_inputs(antigen_coords = antigen_coords, 
                                      relative_immunodominance = c(1,1,1), 
                                      n_epitopes = 3, 
                                      n_antigens = n_antigens,
                                      n_abs_per_serum = 3, 
                                      sigma = .01)

```

## Explore the relationships between map distance and titer distance
```{r}
simple_serum$titer_map %>%
  merge(weighted_avg_distance_map) %>%
  arrange(min_distance) %>%
  ggplot() +
  geom_text(aes(x = titer_distance, y = min_distance, color = dist_variance, label = serum)) +
  geom_abline()+
  ggtitle('Correlation between titer distance and min epitope distance')
```

```{r}
simple_serum$titer_map %>%
  merge(weighted_avg_distance_map) %>%
  arrange(min_distance) %>%
  ggplot() +
  geom_text(aes(x = titer_distance, y = mean_distance, color = dist_variance, label = serum)) +
  geom_abline()+
  ggtitle('Correlation between titer distance and average epitope distance')
```


```{r}
simple_serum$titer_map %>%
  merge(weighted_avg_distance_map) %>%
  arrange(min_distance) %>%
  ggplot() +
  geom_text(aes(x = titer_distance, y = max_distance, color = dist_variance, label = serum)) +
  geom_abline()+
  ggtitle('Correlation between titer distance and max epitope distance')
```



## Plot expectations

```{r}
antigen_coords %>%
  group_by(antigen) %>%
  mutate(mean_c1 = mean(c1),
         mean_c2 = mean(c2)) %>%
  arrange(antigen) %>%
  ggplot() +
#  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), alpha = .2)+
  geom_point(aes(x = mean_c1, y = mean_c2, color = antigen), pch = 7)+
  ggtitle('Expected mean distance map')
```
```{r}
antigen_coords %>%
  group_by(antigen) %>%
  mutate(weighted_c1 = sum(c1*weights),
         weighted_c2 = sum(c2*weights)) %>%
  arrange(antigen) %>%
  ggplot() +
#  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), alpha = .2)+
  geom_point(aes(x = weighted_c1, y = weighted_c2, color = antigen), pch = 7)+
  ggtitle('Expected weighted mean distance map')
```


```{r}
min_distance_df <- epitope_distances %>%
  filter(antigen != serum) %>%
  group_by(antigen, serum) %>%
  filter(distance == min(distance)) %>%
  merge(antigen_coords, by = c('epitope', 'antigen'), all.x = T, all.y = F) %>%
  rename(c1_ag = c1, 
         c2_ag = c2) %>%
  merge(antigen_coords %>% rename(serum = antigen),
        by = c('epitope', 'serum'), all.x = T, all.y = F) %>%
  rename(c1_ab = c1, 
         c2_ab = c2)

all_distance_df <- epitope_distances %>%
   merge(antigen_coords, by = c('epitope', 'antigen'), all.x = T, all.y = F) %>%
  rename(c1_ag = c1, 
         c2_ag = c2) %>%
  merge(antigen_coords %>% rename(serum = antigen),
        by = c('epitope', 'serum'), all.x = T, all.y = F) %>%
  rename(c1_ab = c1, 
         c2_ab = c2)
  

ggplot() +
  geom_text(aes(x = c1, y = c2, color = antigen, label = epitope), data = antigen_coords)+
  geom_text(aes(x = c1, y = c2, color = serum, label = epitope), data = antigen_coords %>% rename(serum = antigen) )+
  geom_segment(aes(x = c1_ag, y = c2_ag, xend = c1_ab, yend = c2_ab, color = as.factor(serum)), data = all_distance_df)+
  geom_segment(aes(x = c1_ag, y = c2_ag, xend = c1_ab, yend = c2_ab), data = min_distance_df)+
  facet_grid(antigen~serum, labeller = label_both) +
  ggtitle('Expected min distance map')
```



```{r}
antigen_coords %>%
  group_by(antigen) %>%
  mutate(mean_c1 = mean(c1),
         mean_c2 = mean(c2)) %>%
  arrange(antigen) %>%
  ggplot() +
#  geom_point(aes(x = c1, y = c2, color = antigen, shape = epitope)) +
  geom_polygon(aes(x = c1, y = c2, fill = antigen), alpha = .2)+
  geom_point(aes(x = mean_c1, y = mean_c2, color = antigen), pch = 7)+
  ggtitle('Expected mean distance map')
```



## Generate sera and titer inputs


```{r}
even_immunodominance_inputs <- generate_ferret_inputs(antigen_coords, 
                                                      relative_immunodominance = c(1, 1, 1), 
                                                      n_epitopes = 3, 
                                                      n_antigens = 5, 
                                                      n_abs_per_serum = 10, 
                                                      sigma = .01)
```





```{r}
E1_complete_dominance_inputs <- generate_ferret_inputs(antigen_coords, 
                                                      relative_immunodominance = c(3, 0, 0), 
                                                      n_epitopes = 3, 
                                                      n_antigens = 5, 
                                                      n_abs_per_serum = 10, 
                                                      sigma = .01)
```

```{r}
E2_complete_dominance_inputs <- generate_ferret_inputs(antigen_coords, 
                                                      relative_immunodominance = c(0, 3, 0), 
                                                      n_epitopes = 3, 
                                                      n_antigens = 5, 
                                                      n_abs_per_serum = 10, 
                                                      sigma = .01)
```

```{r}
E3_complete_dominance_inputs <- generate_ferret_inputs(antigen_coords, 
                                                      relative_immunodominance = c(0, 0, 3), 
                                                      n_epitopes = 3, 
                                                      n_antigens = 5, 
                                                      n_abs_per_serum = 10, 
                                                      sigma = .01)
```

## Save inputs
```{r}
write_rds(even_immunodominance_inputs, paste0(outdir, '/even_immunodominance_inputs.rds'))
write_rds(skewed_immunodominance_inputs, paste0(outdir, '/skewed_immunodominance_inputs.rds'))
write_rds(E1_complete_dominance_inputs, paste0(outdir, '/E1_complete_dominance_inputs.rds'))
write_rds(E2_complete_dominance_inputs, paste0(outdir, '/E2_complete_dominance_inputs.rds'))
write_rds(E3_complete_dominance_inputs, paste0(outdir, '/E3_complete_dominance_inputs.rds'))
```
